A mechanistic theory of planning in prefrontal cortex

Version 1

Abstract

Planning is critical for adaptive behaviour in a changing world, because it lets us anticipate the future and adjust our actions accordingly. While prefrontal cortex is crucial for this process, it remains unknown how planning is implemented in neural circuits. Prefrontal representations were recently discovered in simpler sequence memory tasks, where different populations of neurons represent different future time points. We demonstrate that combining such representations with the ubiquitous principle of neural attractor dynamics allows circuits to solve much richer problems including planning. This is achieved by embedding the environment structure directly in synaptic connections to implement an attractor network that infers desirable futures. The resulting ‘spacetime attractor’ excels at planning in challenging tasks known to depend on prefrontal cortex. Recurrent neural networks trained by gradient descent on such tasks learn a solution that precisely recapitulates the spacetime attractor – in representation, in dynamics, and in connectivity. Analyses of networks trained across different environment structures reveal a generalisation mechanism that rapidly reconfigures the world model used for planning, without the need for synaptic plasticity. The spacetime attractor is a testable mechanistic theory of planning. If true, it would provide a path towards detailed mechanistic understanding of how prefrontal cortex structures adaptive behaviour.

Introduction

While different cortical areas support different functions, common computational principles are shared across many areas. For functions as disparate as sensory processing, spatial reasoning, and language comprehension, features of the environment are inferred from partial information. To do so, structural knowledge about the world must be embedded in synaptic connections (Ko et al., 2011; Burak and Fiete, 2009; Iacaruso et al., 2017; Turner-Evans et al., 2020). This constrains neural circuits to represent meaningful interpretations of the environment, and inputs select between these interpretations. It is not known whether similar principles generalise to complex prospective behaviours, such as planning extended action sequences to achieve a distant goal. Recent recordings from prefrontal cortex give an important clue. When mice or monkeys must execute a sequence of actions, neurons represent the entire sequence concurrently (El-Gaby et al., 2023; Xie et al., 2022). Different neuronal populations encode different steps of the future behaviour. If such a representation could be inferred from inputs indicating goals and constraints, the network could plan the future. Excitingly, this solution would use algorithmic principles similar to those known to infer features of the present in other cortical areas.

This paper has four overlapping aims. (i) To develop a detailed circuit model that infers explicit representations of the future from partial inputs. (ii) To discern the principles of synaptic connectivity that enable such a model to solve complex problems including planning. (iii) To understand when and why this algorithm succeeds while simpler circuit models fail. (iv) To explore how it relates to prefrontal representations, connectivity, and function.

Sequence representations in PFC

Recent work has uncovered PFC representations underlying sequence working memory (Xie et al., 2022; El-Gaby et al., 2023; Whittington et al., 2023; Botvinick and Plaut, 2006). Separate neural ‘subspaces’, or groups of neurons, represent the expected or desired state of the world at different steps along the sequence of future behaviour. In other words, some neural subspace ‘A’ represents the present, while subspace ‘B’ represents the immediate future, and subspace ‘C’ the more distant future (Figure 1A). Critically, these subspaces are active simultaneously. Together, they instantaneously represent the entire behavioural sequence. Importantly, El-Gaby et al. (2023) showed that the PFC subspaces representing different steps of the future are not independent. Neuronal correlations reflect the structure of the task being performed, even during rest.

Background.

(A) Recent work has identified explicit sequence representations in prefrontal cortex during working memory (Xie et al., 2022; El-Gaby et al., 2023). When an animal has to execute a behavioural sequence (left), individual neurons represent conjunctions of location and sequence element (centre). Separate populations (planes) therefore represent the expected location at different times in the future. The entire sequence is represented concurrently by the simultaneous firing of different neurons that encode the expected location at each time in the future (right). (B) An example V1 cell fires when its receptive field is aligned with an inferred line (blue), but not for a control stimulus with no inference (orange). Such visual inference is mediated by structural priors embedded in the circuit connectivity, where neurons representing consistent visual features excite each other (Iacaruso et al., 2017; Shin et al., 2023). Figure adapted from Lee and Nguyen (2001). (C) Structural knowledge is embedded in the synapses of the head direction system (centre; Turner-Evans et al., 2020), which constrains the network to represent single angles (Kim et al., 2017). Visual and proprioceptive inputs (left) determine which angle should be represented (right). We suggest that a mechanism like (B) and (C) infers the representation in (A). (D) Prefrontal cortex is particularly important in dynamic environments. Stimulus-response associations and repeated choices are robust to prefrontal lesions (i; ii), and acortical mice can solve spatial memory tasks (iii; Zheng et al., 2024). However, PFC is needed for reversal learning (iv; Walton et al., 2010) and when different goals are important, or ‘rewarding’, at different times (v; Shallice and Burgess, 1991). In multiplayer board games (vi), different resources are valuable at different stages, and opponents can dynamically change their strategy.

Planning by inferring the future

The correlation structure observed in PFC resembles other attractor circuits known to infer the current state of the world, such as the instantaneous visual input or head direction of an animal (Figure 1B-C; Ko et al., 2011; Chaudhuri et al., 2019). If a similar inference process acted on the future representation in PFC, it would extend its function beyond sequence memory to situations where entire action sequences are inferred from partial cues. Planning could be solved by inferring sequences of desired actions from a set of goals in a network implementation of ‘planning as inference’ (Botvinick and Toussaint, 2012; Levine, 2018). The resulting algorithm would naturally cope with dynamic environments, because it represents each time in the future separately. This is intriguing because PFC is particularly important for problems that require the correct behaviour to be expressed at an appropriate time (Figure 1D; Shallice and Burgess, 1991; Volle et al., 2011). Planning via attractor dynamics is also consistent with winner-take-all dynamics identified in frontal cortex during non-sequential behaviours (Ruff et al., 2025; Inagaki et al., 2019). Such planning as inference differs from most planning algorithms studied in cognitive science and machine learning, which often rely on sequential search (Callaway et al., 2022; Schrittwieser et al., 2020). Search is easily adapted to new environments, but it is slow at decision time. In familiar environments where the structure is embedded in cortical connections, attractor dynamics provide a complementary mechanism for rapid evaluation of many possible futures in parallel.

A mechanistic theory

In this paper, we show that known features of PFC representations and connectivity are sufficient to implement a powerful planning algorithm with minimal additional assumptions. The resulting ‘spacetime attractor’ (STA) instantiates an explicit world model in the synaptic connections between neurons, which allows it to plan by inferring optimal future trajectories. This algorithm resembles other cortical circuits known to infer features of the present, thereby unifying our understanding of PFC with the rest of cortex. The spacetime attractor excels at dynamic problems with changing reward and transition structures, which PFC is critical for and existing mechanistic models cannot solve. RNNs trained on dynamic tasks implement a spacetime attractor in their dynamics, suggesting that it is an efficient solution. Our findings provide a precise mechanistic theory of adaptive behaviour that reconciles prior work on PFC representations with other known cortical computations and deficits from lesions.

Results

An attractor network in space and time

We now introduce the spacetime attractor in more detail. We first review ring and grid attractors, which illustrate how simple circuit motifs can guide inference from partial information. We then explain how the spacetime attractor uses similar principles to infer future behaviour. Neurons are assumed to encode single environment features, but similar ideas apply when individual neurons represent combinations of variables (Clark et al., 2025).

Ring attractors

Ring attractors are neural networks that can only encode circular variables such as angles (Zhang, 1996; Ben-Yishai et al., 1995). Different neurons have different ‘preferred orientations’, and the network dynamics have a set of ‘fixed points’ (stable activity patterns) that encode a particular angle. At these fixed points, neurons with a preferred angle near the encoded angle are active, and the remaining neurons are silent (Figure 2A, bottom left). This is achieved by a network where neurons are mutually excitatory if they have similar preferred orientations, and inhibitory otherwise (Figure 2A, top left). The network then infers which angle is most compatible with noisy inputs such as vision and proprioception. This property underlies the ability of ring attractors to integrate angular velocity in the head direction system (Turner-Evans et al., 2017; Skaggs et al., 1994).

The spacetime attractor.

(A) Left; in a one-dimensional ring attractor, neurons representing similar directions excite each other (top, red), and different directions inhibit each other (blue). Connections are shown for the green example cell. Stable network states (fixed points) have neurons at a particular encoded angle being most active (bottom). Right; grid cells (bottom) emerge from a two-dimensional attractor network (top), where cells with similar preferred location excite each other (red) and intermediate distances inhibit each other (blue). (B) The spacetime attractor is a three-dimensional generalisation, where neurons have both a preferred location (planes) and delay (horizontal axis). This resembles the PFC representation in Figure 1A. Cells that represent adjacent locations in both space and time excite each other (top, red), and different locations at the same time inhibit each other (blue). Given inputs indicating the current location (mouse) and future reward (cheese), the fixed points represent reward-maximising paths through space and time (bottom). While this example has a single static goal, reward inputs can also differ between subspaces if the reward is dynamic. (C) Dynamics of the spacetime attractor. Initial activity is diffuse (left), followed by convergence to a stable representation of a plan (centre). At convergence, a policy is read out from the subspace that represents the next location along the desired trajectory. When the agent moves to the next state, neural activity updates to represent the remaining plan-to-go (right; El-Gaby et al., 2023).

Grid attractors

Grid attractors instead encode position in two-dimensional space. This is achieved by neurons that are mutually excitatory if they represent nearby locations in space and inhibitory if they represent more distant locations (Figure 2A, top right; Fuhs and Touretzky, 2006; Burak and Fiete, 2009). The fixed points of the network are hexagonal patterns of activity that resemble canonical grid cells (Figure 2A, bottom right; Hafting et al., 2005), and the inputs determine which location is represented at a given moment in time. These properties underlie the ability of grid cells to perform path integration (Burak and Fiete, 2009).

Spacetime attractors

Inspired by these known attractor networks in the brain, we propose the existence of a spacetime attractor in prefrontal cortex. The STA is a network that infers explicit representations of the future, and its fixed points therefore have to be entire paths through space and time (Figure 2B, bottom). Given such fixed points, any input would automatically be converted into a representation of future behaviour – an inferred plan. Similar to the ring and grid attractors, this is achieved by connecting STA neurons such that consistent states excite each other and inconsistent states inhibit each other. In the spacetime attractor, each neuron has both a preferred location and a preferred delay δ, which determines how far into the future its tuning curve is defined. If a neuron has δ = 0, it will fire when the agent is currently at the preferred location of that neuron. If a neuron has δ = 3, it will fire when the agent expects to be at the preferred location after three actions. In such a network, consistent states are those that can be part of a single trajectory, while inconsistent states cannot be part of the same trajectory. The connectivity between neurons in subspaces with preferred delays δ and δ +1 should therefore correspond to the structure – more specifically the adjacency matrix – of the environment (Figure 2B, top).

Reward input selects the future

By embedding a world model in its connections, the spacetime attractor creates fixed points that are future trajectories through space and time. It can then use reward information from the environment to bias the representation towards trajectories that represent desirable futures. This resembles how visual and proprioceptive inputs to a ring attractor bias its representation towards particular orientations (Kim et al., 2019). Reward information must similarly be an input to the STA to enable fast adaptation to changing rewards without rewiring the synaptic connections. Importantly, the STA can accommodate time-varying reward structures known to engage PFC (Figure 1D; Shallice and Burgess, 1991; Volle et al., 2011; Carlesimo et al., 2014). This is possible because different neural populations that represent different times in the future can receive different inputs. Given a restaurant booking in 2 hours and a cinema ticket in 4 hours, the δ = 2 neurons would receive reward input at the restaurant, and the δ = 4 neurons at the cinema. The network dynamics would then infer a future where the agent is at each location at the appropriate time.

To summarise, the spacetime attractor has four main components: (i) different neural populations represent different times in the future; (ii) the neurons are connected according to the structure of the environment; (iii) the current location is an input to the ‘present’ subspace; and (iv) the reward structure is an input to all future subspaces. The resulting fixed points (Figure 2B, bottom) represent trajectories that maximise cumulative reward. The network dynamics implement a gradual relaxation process, where reward inputs bias the representation in each subspace towards high-reward locations, while inputs from neighbouring subspaces constrain their representations to be consistent. Together, these two processes converge to a stable representation of a coherent future plan (Figure 2C).

The STA guides behaviour

After computing an explicit representation of a plan, it can be used to inform behaviour. In particular, an agent should simply take whichever action leads it to the next location on the inferred trajectory (Figure 2C, middle). The entire representation of the future then needs to move by one ‘action’, so the state that was previously represented in subspace δ is now in subspace δ − 1. The resulting dynamics resemble a ‘conveyor belt’ that allows the STA to execute entire trajectories without recomputing the policy after every action. Such conveyor belt dynamics have been observed experimentally during sequence working memory (El-Gaby et al., 2023). Importantly, the representation remains a fixed point of the spacetime attractor dynamics, because the location and reward inputs are updated due to (i) the movement of the agent, and (ii) the passing of time.

World models for planning

Planning requires access to an internal world model that predicts the consequences of different actions. However, the spacetime attractor uses such a world model differently from many other planning algorithms. Most algorithms apply a world model sequentially to simulate actions one by one. The spacetime attractor instead instantiates multiple copies of the world model explicitly in the synapses between different neural subspaces. This allows the network to simulate many possible futures in parallel. Entorhinal grid cells are also thought to embed a world model in their connectivity (McNaughton et al., 2006), but they do not represent the distant future explicitly (Ouchi and Fujisawa, 2024). Instead, they infer a single location at a time. The spacetime attractor therefore suggests that circuit principles in prefrontal cortex resemble other cortical areas that use structural knowledge to infer features of the world. We propose the major difference to be that PFC instantaneously represents the world at many points in time, which extends known circuit principles to complex planning problems. This also requires knowledge of the reward available in different states, which could be estimated separately by other neural circuits. In this work, we simply assume access to ground-truth rewards.

STAs are flexible planners

Prefrontal cortex is particularly important for tasks that require flexible behaviour in changing environments (Figure 1D; Burgess and Wu, 2013). To understand whether the spacetime attractor is a good model of planning in PFC, we therefore need to (i) study its behaviour and performance in such dynamic tasks, and (ii) characterise when and how it differs from existing models. We will see that the spacetime attractor is well-suited to dynamic PFC-like’ problems, which other mechanistic models struggle to solve.

From static to dynamic tasks

We designed a set of four tasks that vary in how much the reward changes in space and time (Figure 3A). The tasks are all embedded in Euclidean space for simplicity of exposition, but the underlying principles generalise to any environment with known structure. Task 1 is simple navigation towards a static goal that remains constant across trials. In task 2, the static goal changes between trials. Task 3 is navigation to a goal that also moves within each trial, where each location is only rewarded when the goal is there. Task 4 generalises the idea of time-dependent goals to a non-binary reward landscape. Reward magnitudes are sampled independently between -1 and +1 for each location at each time point in each trial. The objective is to maximise cumulative reward over the trial, which requires balancing immediate reward with the potential for future reward (Figure 3A, right). This is reminiscent of the example from Figure 1D, where the restaurant and cinema are desirable at different times. To better understand how the STA solves these tasks, we compare it to two models commonly studied in systems neuroscience. The first is temporal difference (TD) learning, which has neural correlates in striatum (Sutton, 1988; Schultz et al., 1997). The second is the successor representation (SR), which has neural correlates in hippocampus (Dayan, 1993; Stachenfeld et al., 2017). These models assume fixed rewards across trials (TD) or within a trial (SR). Unlike the spacetime attractor, they do not generalise well when the reward changes rapidly.

Model comparisons in static and dynamic tasks.

(A) Example tasks with different degrees of dynamic reward. Colours indicate reward at each location (white to green), and red arrows indicate the optimal policy. In the ‘moving goal’ task, the agent intercepts a target that moves along a different trajectory in each trial (arrow). In the ‘reward landscape’ task, the reward is sampled independently in space and time. The agent has to maximise cumulative reward, and it can be optimal to forgo immediate reward for a later payoff. (B) Representations and performance in the static goal task with a fixed goal. Left: value functions learned by TD and SR agents. Centre: STA representation at convergence, which encodes a path through space and time. The final plane is a max projection that summarises the path. Right: performance of each model in the task (grey: random baseline). (C) As in (B), now for the task where the static goal changes between trials. (D) Representations and performance in the moving goal task. Left: the SR computes a value function that averages reward across time-within-trial. Middle: the STA takes into account the moving goal and computes a path that intercepts it. Right: performance of each agent. (E) Performance in the reward landscape task. All error bars indicate 1 standard deviation across 20 agents and environments (dots).

Simpler models can solve static tasks

TD learners gradually propagate value from rewarded locations to all other locations, and optimal policies are only learned when rewards are constant across both time and trials (Figure 3B-C, orange). The SR agent uses the environment structure and trial-specific reward function to compute values, and it can solve tasks where the goal changes between trials (Figure 3B-C, green). The STA can also solve these tasks, but it does so by inferring an entire spatial trajectory to the goal instead of computing a value function (Figure 3B-C, blue). Because the reward is constant throughout a trial, each subspace receives the same reward input. However, the inferred representation differs across subspaces because it is also constrained to satisfy the current agent location and the transition structure of the world. As we will see, the inference of an explicit spacetime representation by the STA increases its flexibility compared to the amortised state values computed by TD and SR agents.

Only the STA solves dynamic tasks

While the SR agent can adapt to rewards that change across trials, the reward structure still needs to remain constant for the duration of the planning horizon. This is because the SR computes time-averaged occupancy and lacks fine-grained information about when the agent is where. When the reward changes within a trial, the SR is limited to computing average values across time-within-trial, and it fails to intercept the goal at the correct point in space and time (Figure 3D). In contrast to the TD and SR agents, the STA can solve the moving goal task because the input to each subspace is the reward at that specific moment in time. The representation is therefore biased towards coinciding with the goal in both space and time, and the network dynamics relax to a future trajectory that correctly intercepts the goal (Figure 3D). The advantage of the STA is exacerbated in the more dynamic reward landscape task, where very different locations can be rewarded at different times (Figure 3E). These results suggest that different algorithms could contribute differently to decision making. TD learning could drive rapid decisions in environments with stable rewards; the SR would be more flexible when rewards change on intermediate timescales; and the STA could enable adaptive behaviour in environments with rapid changes. Finally, sequential search would facilitate slower planning in novel environments and could be guided by partial plans from a spacetime attractor.

STAs are efficient planners

We have seen that a handcrafted spacetime attractor can solve dynamic problems that simpler algorithms cannot. However, this requires many neurons to explicitly represent the future. We therefore investigated whether superior solutions exist to these ‘PFC-like’ problems by training a recurrent neural network (RNN) to solve them efficiently (Mante et al., 2013; Stroud et al., 2023). We will see that regularised RNNs solve dynamic planning problems with an STA-like algorithm, suggesting that it is efficient. This result also demonstrates that the STA is consistent with a prominent theory that PFC resembles a recurrent meta-learner (Wang et al., 2018). In this view, PFC learns connections over long timescales that implement adaptive behaviour on short timescales through recurrent dynamics. This is exactly how planning happens in the STA, which embeds task structure in the weights and uses recurrent dynamics for rapid decision making with different rewards. We extend the metalearning theory of PFC by showing that an STA is the mechanism implemented in the dynamics of RNNs meta-trained on challenging planning tasks.

The three defining features that allow a spacetime attractor to plan via inference are: (i) it has an explicit representation of the future; (ii) the connectivity embeds a model of the world; and (iii) attractor dynamics infer the future representation. We will see that RNNs trained on the reward landscape task exhibit all of these properties. The RNNs were trained with supervised learning to generate optimal actions from inputs that indicated the current location and trial-specific reward structure (Figure S1; Methods; see Supplementary Note for a discussion of different modelling choices). The reward input was only provided during an initial ‘planning phase’ prior to the ‘execution phase’ of the task, and the loss function also penalised the magnitude of neural firing rates and network weights to encourage an efficient solution.

RNNs learn explicit future representations

Spacetime attractors compute explicit representations of the future (Figure 4A-B). Critically, these representations generalise over trajectories. In other words, the representation in subspace δ depends only on the expected location in δ actions, and not on the rest of the trajectory or the history of the agent (Figure 4A, right). To confirm that the RNN learned such a representation, we trained linear decoders to predict the future from its hidden state. To ensure generalisation across trajectories, we trained the decoders while holding out each ‘cur-rent location’ and computed the test accuracy only from those held out locations. The RNN had an explicit representation of the future during both planning and execution (Figure 4C-D; Figure S2). In a spacetime attractor, the same subspace always encodes location δ actions into the future. The future representation of the RNN also exhibited such ‘conveyor belt’ dynamics (Figure 4E). The neural representation of the RNN is thus consistent with a spacetime attractor.

RNNs learn spacetime representations in the reward landscape task.

(A) An STA infers the entire future during planning and represents the ‘future-to-go’ during execution (left). The representation in each subspace generalises across all trajectories passing through that location in spacetime (right). (B) We compare a spacetime attractor; an RNN trained on the reward landscape task; and an agent that computes an exact value function in space and time. The value-based agent computes an optimal policy from ‘neural activity’ containing (i) the value function, (ii) the agent location, and (iii) the time-within-trial (Methods). (C) Decoding accuracy at the end of planning for: left; agent location at each time in the future. Right; the time at which the agent will be at a given location, plotted as a function of the actual time the location was visited. Decoders were trained in crossvalidation across the current agent location (Methods). (D) Decoding accuracy during execution of location at each time in the past or future. (E) We trained a single decoder to predict location at time 3 from neural activity at time 1 (black circle). The same decoder predicted location at time t + 2 from neural activity at any other t, demonstrating ‘conveyor belt’ dynamics. (F) Performance in the static goal (left), moving goal (centre), and reward landscape (right) tasks for RNNs trained on either task (x-labels; colours). (G) Normalised parameter magnitudes of the three RNNs (left) and average firing rates in the static goal task (right). All error bars indicate 1 standard deviation across 5 RNNs (dots).

Intriguingly, RNNs trained on the simpler static and moving goal tasks did not reliably learn explicit future representations (Figure S3). Additionally, the RNN trained on the reward landscape task could solve these simpler tasks, while RNNs trained on the simpler tasks failed in the reward landscape task (Figure 4F). Alternative solutions therefore exist in the simpler tasks, and these solutions are favored by a pressure to be energetically efficient (Figure 4G). We identify a tradeoff between generality and efficiency, where the PFC-like solution is general but needs more neurons and synapses, while simpler algorithms can solve simpler tasks more efficiently. Unlike other theories of PFC such as value coding, this suggests a reason why PFC occupies such a large cortical territory.

RNNs learn a world model

In a spacetime attractor, subspaces that represent expected future locations are connected according to the environment structure. Testing this prediction in the RNN requires interpretation of the network weights, which is challenging because the functions of individual neurons are unknown. We therefore project the weights into a coordinate system, where the axes are orthogonal directions in neural state space that predict each point in spacetime (Methods). While this does not perfectly recover the true subspaces in the handcrafted STA, it is a good estimate. The projected RNN weights can therefore be interpreted as interactions between representations of different points in spacetime (Figure 5A).

RNNs learn a world model.

(A) The weights of RNNs trained on the reward landscape task are projected into an orthonormal coordinate system with axes that predict different points in spacetime (top). The projected weights are interactions between points in spacetime (bottom). (B) The average recurrent weights between subspaces separated by a single action resemble the environment adjacency matrix (bottom, ‘RNN’). ‘Empirical STA’ is the same analysis performed on approximate subspaces estimated from neural activity in the handcrafted STA. ‘True STA’ indicates weights between the ground truth future-coding subspaces in the handcrafted STA. Green box in ‘True STA’ indicates weights between (i) the single location denoted by a green circle in ‘Prediction’, and (ii) all locations in the subspace indicated by a green square. (C) Input weights to the ‘current’ (δ = 0; left) and a ‘future’ (δ = 2; right) subspace. Consistent with an STA, the current subspace of the RNN receives location input, and future subspaces receive reward information. (D) Average recurrent weights between an example location (green arrow) and other locations at increasing time differences (Δ; planes). Locations in subspaces Δ apart are more strongly connected if they can be reached in Δ actions. Light green circles indicate these Δth order adjacency matrices (Δ = 0 corresponds to the identity). (E) Correlation between (i) the average connectivity between subspaces separated by Δ actions (lines; legend), and (ii) different order adjacency matrices (x-axis). Shading indicates standard deviation across 5 RNNs. (F) Example environment with a high-value path and a lower-value path. (G) Weak stimulation does not affect the spacetime representation of the RNN, but strong stimulation switches it to the lower-value path. (H) Magnitude of representational change over time across stimulation strengths (Methods).

Remarkably, the recurrent weights between adjacent subspaces closely resembled the adjacency matrix of the environment (Figure 5B; Figure S4; mean ±std correlation of 0.91 ±0.07 vs. 0.72 ±0.06 for control environments). The RNN thus learned a world model explicitly in its recurrent weights. Consistent with a spacetime attractor, the δ = 0 subspace received input indicating the current agent location, and subspace δ received input indicating the reward in δ actions (Figure 5C). The RNN also learned weaker connections between more distant subspaces, and these connections reflected the environment structure (Figure 5D-E; Supplementary Note). In summary, the RNN learned all of the structural components that allow a spacetime attractor to infer future actions by integrating expected reward across time using a world model.

RNNs learn attractor dynamics

In a spacetime attractor, the structured connections give rise to attractor dynamics with fixed points corresponding to explicit representations of desirable futures. To demonstrate similar attractor dynamics in the trained RNN, we artificially perturbed its neural activity at the end of planning. This resembles how attractor dynamics have been demonstrated in biological circuits (Inagaki et al., 2019; Kim et al., 2017; Vinograd et al., 2024). An attractor network should be robust to small perturbations, and the representation should be more sensitive to perturbations towards other attractor states than perturbations along random directions in neural state space. We analysed the RNN in a setting with two ‘good’ paths that were close in value (Figure 5F). Like the hand-crafted STA, the RNN initially converged to a stable representation of the better path, which persisted in the presence of weak perturbations. Stronger perturbation of a single location on the alternate path caused a discrete switch to a representation of this entire path, including in subspaces not directly perturbed (Figure 5G). Perturbations of similar magnitude in random control directions had little effect (Figure 5G, grey). Unlike the handcrafted STA, the RNN representation relaxed back to the original fixed point after removing the perturbation (Figure 5H; Figure S5), which suggests that the RNN learned an attractor landscape that is less susceptible to local minima.

Together, our analyses show that RNNs trained on a dynamic planning task learn to closely approximate a spacetime attractor. This was also true across variations in model architecture (Methods; Figure S6; Figure S7). Finally, RNNs with too few hidden units to learn a spacetime attractor failed to solve the task (Figure S8), suggesting that other solutions are not readily learned by gradient descent. The spacetime attractor is thus an efficient solution to dynamic planning problems, and it is consistent with a prominent theory that PFC can be understood as a recurrent meta-learner.

STAs can generalise across environments

Hardwiring the transition structure of the environment into the synapses of a spacetime attractor seems prohibitive, because it might prevent the network from generalising across different structures (Figure 6A). Here, we show that is not the case. RNNs trained in environments with changing structure still learn spacetime attractors. The network dynamics are adapted to each environment through input-mediated gating of a generic scaffold. This is a simple mechanism for rapid adaptation that preserves the ability to plan under changing rewards.

Spacetime attractors can adapt to changing structure.

(A) Changing structure requires adaptation. (B) Performance of RNNs trained on the reward landscape task in a single maze or with a different structure on every trial, when evaluated in either a single maze (top) or across changing mazes (bottom). (C) Two example mazes (left) and their corresponding adjacency matrices (centre). The effective connectivity between adjacent subspaces in a single RNN (right) reflects the structure of each maze. (D) Average correlation between subspace connectivity in a maze and the structure of either the same (blue) or a random (grey) maze. (E) The spacetime representation is more similar across distinct trials from the same maze than trials from different mazes (left). By using slightly different subspaces in different environments, the network can match their connectivity to the environment structure (right; orange vs. blue). (F) Putative mechanism for structural generalisation. Instead of representing each future location, neurons encode expected future transitions (black arrows in example state). Each transition (green) connects to transitions that can follow or precede it (red arrows). Structural input to PFC inhibits transitions that are not available in a given environment (blue), preventing planning between states that would otherwise be connected (light red arrows). (G) Effective connectivity between directions in neural state space that encode consistent’ consecutive future transitions (and ; green to red), ‘adjacent’ transitions (green to blue), or any other transitions (green to grey). (H) Projection of the input from wall wij onto representations of future transitions through the wall (τij; dark blue), transitions to other adjacent states (τik; light blue), or any other transitions (grey). All error bars indicate 1 standard deviation across 5 RNNs (dots).

Representations adapt to the environment

To study planning in environments with changing structure, we trained RNNs on a variant of the reward landscape task, where the transition structure was different on every trial (Methods). The structure was provided to the RNN as a binary input that indicated whether pairs of otherwise adjacent states were separated by a wall. Importantly, the network had to adapt to each new environment through its dynamics while keeping the weights fixed across environments (Wang et al., 2018; Jensen et al., 2024). The generalising RNNs performed nearly as well as networks trained in a single environment (Figure 6B), raising the question of how generalisation is achieved. The neural representation resembled a spacetime attractor in any single environment (Figure S9B). However, the effective connectivity between subspaces changed between environments. Remarkably, the connectivity reflected the transition structure of whichever environment the agent was currently in (Figure 6C-D). This indicates a general solution that allows recurrent networks to use attractor dynamics for planning in any similar environment, without requiring synaptic plasticity.

A model of generalisation

We can understand how the network adapts to the environment at both the ‘population level’ and at a mechanistic level. At a population level, the directions in neural state space that represent each point in spacetime were more similar in different trials from the same environment than across trials from different environments (Figure 6E, left). This change allows the network to align its representation with the appropriate components of the connectivity matrix (Figure 6E, right). How is this achieved mechanistically? We hypothesised that the network explicitly represents future transitions rather than future locations. In other words, there are directions in neural state space that represent ‘in δ actions from now, move from state si to state sj’, which we denote . This is different from the vanilla spacetime attractor, which has a representation of future locations that is invariant to the subsequent state.

While it requires more neurons, explicitly representing future transitions allows the network to use input-mediated inhibition to adaptively gate off transitions that are not available in any particular environment. The learned structure of the remaining transitions is then used for planning (Figure 6F). Importantly, the scaffold only needs to include transitions that are possible in at least one environment. This model makes three predictions that we verified: (i) the network explicitly represents future transitions (Figure S9C), (ii) the representations of future transitions are connected by a world model (Figure 6G), and (iii) input indicating a wall between si and sj specifically inhibits representations of future transitions between those states (Figure 6H). These analyses show that the spacetime attractor is an efficient planning algorithm in structurally changing environments, and they suggest a general mechanism for adaptive behaviour.

Discussion

We have taken inspiration from known cortical attractor dynamics and recent findings on prefrontal sequence representations (Figure 1) to develop a new theory of planning. This spacetime attractor model uses different neural subspaces to represent the expected location of an agent at different times in the future. The subspaces are connected according to the structure of the environment, which allows the network to infer optimal future trajectories (Figure 2). The spacetime attractor can solve problems with dynamic rewards, which PFC is important for and simpler algorithms struggle with (Figure 3). It solves these problems using internal representations that closely resemble prefrontal representations during sequence working memory, thereby unifying working memory and planning in PFC. RNNs trained to solve dynamic planning tasks learn to implement the spacetime attractor algorithm in their internal dynamics, suggesting that it is an efficient solution (Figure 4; Figure 5). Finally, spacetime attractors can generalise across environments with different transition structures through rapid gating of a general scaffold to reflect each particular environment (Figure 6).

Experimental predictions

The spacetime attractor is inspired by data and makes precise predictions that can be tested in future experiments:

After planning a behavioural sequence, different subspaces of PFC activity should represent distinct steps of the plan.

Optogenetic activation of neurons in a future subspace should bias representations and behaviour towards the stimulated state at a delay corresponding to the stimulated subspace.

The effective connectivity between subspaces should reflect the structure of the environment. This could be tested in noise correlations across neurons or by explicit holographic stimulation.

Different patterns of neurons should represent the future in environments with different transition structures.

The neurons active in each environment should be connected according to its structure. Inputs that mediate this gating could come through sensory cortex in observable environments or hippocampus when the structure is learned.

The first two predictions would also be true of independent subspaces used to store sequence memories (Xie et al., 2022; El-Gaby et al., 2023), while the last three are unique predictions of a structured network that plans behavioural sequences.

Decision making across the brain

Our comparisons of different models across different tasks suggest complementary contributions of different brain regions to decision making. Striatum is thought to implement temporal difference learning (Schultz et al., 1997), which facilitates rapid responses in stable environments. Hippocampus embeds the structure of the world, but it is thought to either average representations over future timesteps in a successor representation (Stachenfeld et al., 2017) or represent one state at a time in a sequence (George et al., 2021; Whittington et al., 2020; Jensen et al., 2024). This provides an efficient solution to problems with consistent spatial structure and rewards that change on intermediate timescales. A spacetime attractor in frontal cortex facilitates adaptive behaviour in dynamic environments with a familiar structural scaffold that has been embedded in PFC. Finally, explicit search would facilitate slower planning in novel environments and could be guided by partial plans from a spacetime attractor. While spacetime attractors are particularly useful in dynamic environments, they can also solve simpler problems. It is therefore possible that animals have developed a spacetime attractor because they sometimes need it for planning, and then reuse it in simpler tasks. However, it is also plausible that many laboratory behaviours instead engage simpler algorithms in the basal ganglia or hippocampus. If that is the case, richer spacetime problems may be needed to query the representations used for planning in prefrontal cortex.

Interactions with other planning algorithms

Planning via inference in a spacetime attractor differs from most models of planning in cognitive science. In particular, the spacetime attractor posits that planning can happen via recognition, where the sequence of steps needed to reach some desired state is directly inferred. In contrast, many studies of human planning focus on explicit search, where different paths are simulated and evaluated sequentially. We suggest that these two processes coexist, with spacetime attractor dynamics facilitating rapid planning as inference in familiar environments, where the structure has been embedded in prefrontal connections. The STA could also help focus explicit search towards putative high-value paths and evaluate the utility of different paths. This could happen through interactions with hippocampal activity sequences that have been proposed to facilitate planning (either replays or theta sequences; Foster, 2017; Widloski and Foster, 2022; Jensen et al., 2024). In particular, Jensen et al. (2024) suggest that PFC biases which sequences are replayed in hippocampus. PFC would then update its representation to make high-reward sequences more likely in an iterative policy improvement process. Hippocampal replay has also been proposed to implement a ‘DYNA’ algorithm, where past experience is used to learn a model-free policy during rest or sleep (Mattar and Daw, 2018). This is complementary to both the spacetime attractor and explicit search. DYNA facilitates rapid decision making when the optimal policy remains stable across time, while decision-time planning allows adaptation to changing environments.

Learning a spacetime attractor

In the hand-crafted spacetime attractor, we embedded copies of the environment adjacency matrix in the connections between every pair of consecutive subspaces. The RNN analyses show that such structure can be learned from repeated experience. However, learning copies of the same parameters independently for each pair of subspaces is inefficient. It would be more efficient to store a cache of experienced trajectories that can be used to learn all of the parameters. Hippocampal replay has been proposed to build cognitive maps via such interactions with prefrontal cortex (Bakermans et al., 2023; Ou et al., 2025). In particular, Ou et al. (2025) suggest that hippocampal replay consolidates structural information from hippocampus into cortex. Replay from hippocampus to PFC during sleep or rest could therefore provide a mechanism for learning all of the STA parameters from the same data.

The state space of planning

We have assumed that planning happens at the level of individual locations in the environment. However, humans often plan in more abstract spaces, which improves efficiency by reducing the required planning depth. We also plan hierarchically by first computing an abstract plan that can be refined in increasing levels of detail (Eckstein and Collins, 2020). Multiple spacetime attractors operating at different levels of abstraction can be coupled to implement such a hierarchical planner. One ‘abstract’ STA would compute a high-level plan. A second STA would treat the next abstract state as a goal and compute a detailed plan to get there. Stacking spacetime attractors in this way allows inference of plans that are exponentially long in the depth of the hierarchy, and therefore the number of neurons, in contrast to the linear scaling of a non-hierarchical STA. However, learning appropriate abstractions and embedding them in an STA remains an unsolved challenge.

Bidirectional environment interactions

We have focused on environments that evolve independently of the agent. The rewards and structure of the environment can therefore be estimated up front and provided as inputs to a spacetime attractor in PFC. However, many problems involve structure that depends on the behaviour of an agent. One example is ‘key-door’ problems, where moving through a locked door becomes possible only after picking up the key. We posit that such problems can be solved by structural gating of a learned scaffold, similar to the STA that generalises across environment structures (Figure 6). The structural gating would no longer be an external input, but instead a function of the spacetime representation itself. For example, activating a cell that represents ‘being at the key’ could disinhibit ‘transitioning through the door’ at a later time.

Another important example of bidirectional interactions is social behaviours, where agents can change their plans in response to each other. A system of two spacetime attractors could simulate the joint dynamics of two such agents. The first STA ‘A’ infers future behaviour for agent A. The reward input would be the output of a different STA ‘B’ that predicts the behaviour of another agent. The reward for STA B would itself be an output of STA A, which couples the predicted behaviour of the two agents. The combined system relaxes to a fixed point where the behaviour of A is optimal given the predicted behaviour of B and vice versa – a putative neural implementation of ‘theory of mind’.

Outlook

We have developed a new theoretical framework for planning in prefrontal cortex. It builds on recently discovered prefrontal working memory representations and known attractor dynamics in other neural circuits. The spacetime attractor extends these principles to prospective behaviours in complex and changing environments – a setting that has previously eluded mechanistic circuit models. Existing data does not allow us to test these ideas explicitly. Instead, we have provided a series of concrete predictions for future experiments. We hope this will inspire new work in both experimental and computational neuroscience.

Methods

Tasks

We used three different classes of tasks to train and compare models in this work – the ‘static goal’ task, the moving goal’ task, and the ‘reward landscape’ task. All tasks had agents navigate mazes on a 4×4 grid, with walls preventing transitions between some pairs of otherwise adjacent states. The wall configurations defining the mazes were sampled as described by Jensen et al. (2024). For most analyses, the wall configuration remained fixed across all trials, and only the reward function changed. For the analyses in Figure 6, the wall configuration changed across trials, and the agent had to adapt to the new transition structure through its recurrent network dynamics. In all tasks, the agent location was sampled randomly at the beginning of each trial. At the beginning of each trial, the agent location and reward function were ‘frozen’ for 5-7 (randomly sampled on each trial) iterations of the environment. This constituted an initial ‘planning’ phase that was followed by an ‘execution’ phase, where (i) the agent took actions that changed its location, and (ii) the reward function updated as described in more detail below. For all analyses, the full reward function was provided to the agent during the initial planning period. For the handcrafted STA and the RNN in Figure S6, the input provided reward information that was relative in time during execution – it indicated what the reward would be at a given location in δ steps rather than at time t. The reward input was zero for all δ that corresponded to time points beyond the end of the trial. For all other RNN analyses, the reward input was set to zero during execution.

Moving goal task

In this task, a full goal ‘trajectory’ was sampled on each trial. The goal trajectory was sampled as a random walk that could only turn around if it reached a dead end. The start location of the agent was restricted to not coincide with the start location of the goal. The trial terminated when the agent was at the same location as the goal at a given moment in time, or after a maximum of six actions. The reward input to the agent, R ∈ ℝT×N, was a matrix indicating the location of the goal at every point in the future. The reward input consisted of ‘+0.6’ for the goal location and ‘-0.6’ for all other locations at each moment in time. In other words, Rδi was ‘+0.6’ if the goal would be at location si δ steps into the future, and −0.6 otherwise.

Static goal task

This task was identical to the moving goal task, except that the goal remained stationary for the duration of each trial. The static goal task was implement in two different variants – one where the goal remained fixed across all trials, and one where the goal was resampled at random in every new trial. The handcrafted models were evaluated in both of these tasks, while the RNNs were only trained with a goal that changed between trials.

Reward landscape task

In this task, every element R(t, s) of the reward function ℛ was sampled uniformly and independently between -1 and +1. The complete future reward structure was provided as an input to the agent as described above. Every trial finished when the agent had taken six actions.

Quantification of performance

For most performance comparisons, we computed the probability of choosing the optimal first action in each trial. We define the optimal first action as the first action of the trajectory that maximises cumulative reward over the entire trial. We use this metric rather than the actual cumulative reward for two reasons. (i) We are interested in the process of planning, whereby an agent balances immediate and long-term reward. This is most challenging for early actions, while the greedy policy is optimal for the last action. (ii) The probability of choosing an optimal action is readily interpretable as a number between 0 and 1. All results were qualitatively similar if we instead used the probability of choosing an optimal action averaged over the entire trial as a performance metric, or if we used the average empirical reward.

Handcrafted models

Here we provide an overview of the handcrafted STA, TD, and SR agents used in the paper.

Spacetime attractor model

The spacetime attractor is a recurrent neural network with an exponential nonlinearity and within-subspace normalisation. We define zδi as the ‘potential’ and rδi as the ‘firing rate’ of a neuron that represents ‘being at location si in δ actions’. The spacetime attractor then has the following network dynamics:

Here, τ = 50 iterations is the time constant of the dynamics, ϵ = −100 is a small constant that thresholds the maximum inhibition between and within subspaces, and η∼ N (0, σ = 0.1) is white noise added to the network dynamics. Aij are weights corresponding to the adjacency matrix, which indicates whether si can be reached from sj in one action. is an input that reflects the normalised reward available at location si in δ actions. In particular, , where Rδi is the trial-specific reward described above, and we set β = 9. To provide an input indicating the current agent location, we set R0i = 20 if si is the current location and 0 otherwise. Since the activity is explicitly normalised within each subspace, the resulting rδi can be interpreted as a distribution over desired locations in δ actions. When simulating the behaviour of the STA, we took the policy of the agent to be greedy in the representation in subspace δ = 1 of locations accessible from the current state, . We also added a small amount of noise ηij ∼ U (−0.025, −0.015) to each element of the weight matrix instead of using the exact adjacency matrix. Noise was added to ensure robustness, and it was restricted to be negative by adding a bias term that prevents representations from teleporting’ between locations (Supplementary Note). An additional ‘feedforward component’ in the form of the identity matrix was added to the connectivity between subspace δ and δ− 1 during the first 100 network iterations (2 time constants) after every action. This is inspired by the ‘update neurons’ of the fruit fly head direction circuit (Turner-Evans et al., 2017) and stabilises the ‘conveyor belt’ dynamics, but it is not necessary for any of the main results in the paper (Supplementary Note). We ran the STA dynamics for 400 network iterations before each action to ensure convergence to a stable representation.

Baselines

Here we provide details of the temporal difference and successor representation baselines that we compare the spacetime attractor to. For further details, we refer to Jensen (2023). Common to both of these frameworks is that they explicitly estimate the ‘value function’ under some policy π:

Here, indicates an expectation taken over trajectories τ resulting from the agent following π. The TD’ and ‘SR’ heatmaps in Figure 3 show the computed value functions. In both cases, we take the policy of the agent to be greedy in the value function evaluated at all locations accessible from the current state.

Temporal difference learning

We implement vanilla temporal difference learning, which computes a value function by iteratively applying the update

We set the temporal discount factor to γ = 1, since we are interested in maximising the non-discounted cumulative reward. We allowed the TD agent to interact with the environment for 4,000 trials with a learning rate of α = 0.05 before analysing its performance.

Successor representation

In the successor representation formalism, the value function is decomposed as

Here, r is a vector of the average reward associated with each state, and is a vector of the expected discounted future occupancy of state s′ if the agent starts in state s and follows policy π:

The full matrix Mπ, constructed from stacking the corresponding to all states s, is denoted the ‘successor matrix’, and it allows us to write down a vector of expected rewards from all states as

M π can be learned using a TD-like algorithm as above. However, since the STA algorithm also has explicit access to the transition structure of the environment, we simply computed the exact successor matrix as the geometric series M π = I + γTπ + γ2(T π)2 + … = (I − γT π)−1 with γ = 0.95. When computing the successor matrix, we take π to be the diffusion policy, and T π is therefore the diffusion matrix. This is similar to how the adjacency matrix serves as a global structural prior in the STA.

Spacetime value agent

In Figure 4, we compare representations to an agent that computes a value function in a state space consisting of space and time. This agent uses dynamic programming to compute the value of being at location s at time t under an optimal policy for every combination of s and t:

At time t, the optimal policy is then greedy in the value of accessible locations at time t + 1. For the decoding analyses in Figure 4, we took the ‘neural representation’ to be the concatenation of (i) the flattened value function vf ∈ ℝTN, (ii) a one-hot representation of the current location, and (iii) a one-hot representation of the current time-within-trial. These three quantities are sufficient to compute an optimal policy.

Recurrent neural networks

In this section, we provide details of how recurrent neural networks were trained and analysed. All networks were fully connected with Nrec = 800 hidden units (except Figure S8) and ReLU nonlinearities. The network dynamics were given by

Here, z is the network potential, r is the ‘firing rate’, x is the input, θ = {Win, Wrec, Wout, brec, z0, bout} are the network parameters, and η ∼ N (0, σ = 10−3) is Gaussian noise. All simulations used a time constant of τ = 5 iterations. The output policy was defined in global allocentric coordinates as the desired next state, π ∝ ey. In Figure S7, we also analyse a network that produced a policy in ‘local’ coordinates that indicated the desired movement direction. The input consisted of (i) a one-hot representation of the current location in the environment; (ii) a binary representation of the location of all walls (Jensen et al., 2024); and (iii) the reward Rδi as described for the tasks above. Gaussian noise with a standard deviation of σ = 10−3 was added to the inputs before passing them to the RNN.

For all analyses in the main text, the RNN performed 10 network iterations per environment iteration (action). Reward input was only provided during the planning phase and set to 0 during the execution phase. This was done to avoid biasing the execution-time representation towards a ‘relative’ subspace representation by providing reward input in that format (Supplementary Note). For comparison with this RNN, we also trained an RNN with reward input during both the planning and execution phase, which learned qualitatively similar representations and dynamics (Figure S6). For this network, the reward input was given relative to the current time as in the handcrafted STA, and the number of network iterations was randomly sampled between 9 and 11 before each environment iteration.

We optimised all network parameters θ to minimise the loss function

is the set of optimal actions at time t, and indicates an expectation over trajectories induced by the policy of the agent, renormalised over optimal actions. That is, we consider an imitation learning setting where ties between equally optimal actions are broken according to the actual policy of the agent. We used λθ = 2 × 10−7, λr = 10−5, and λe = 10−4 for all analyses. The RNNs were trained using Adam (Kingma and Ba, 2015) with a learning rate of 3 × 10−4 for 200,000 batches of 200 trials (250,000 batches for the RNNs trained in changing mazes).

All results are reported as mean and standard deviation across 5 separate RNNs that were trained from different random seeds and with different environment transition structures.

Analyses

In this section, we describe the analyses used to compare different handcrafted models and trained RNNs.

Decoding analyses

To decode future locations from neural activity (Figure 4C-D), we used scikit-learn (Pedregosa et al., 2011) to train L2-regularised logistic regression models that predicted location at time tL from neural activity at time tN with an inverse regularisation strength of C = 1.0. We performed this analysis in crossvalidation across locations at time tN. In other words, we trained a decoder on data where the agent was in any state to predict all locations at time tL, and we then tested this decoder in trials where the agent was in state s at time tN. We repeated this analysis across all test locations s and averaged performance across the resulting 16 folds. We did this to test whether the RNN had a generalisable representation of future location, rather than an encoding of e.g. current location and neighboring values.

In Figure 4C (left), we plot the performance of decoders trained on tN = −1 to predict location at any tL ≥ 1. In Figure 4D, we trained decoders on every pair of tN ∈ [0, 5] and tL ∈ [0, 5]. We then averaged the performance across every ‘delay’ tL − tN.

To investigate the generalisation of decoders in Figure 4E, we trained a single decoder using neural activity at time tN = 1 to predict location at time tL = 3. We then applied the same decoder to neural activity at all times and quantified how well it predicted location at all times . This analysis was again performed in crossvalidation. A separate decoder was trained while holding out each location s1 = s, and then tested only on trials where . Performance was averaged across all held-out locations and across 5 independently trained RNNs. Figure 4E shows the time at which the average predictive performance was highest for each . Figure S2 shows the full generalisation behaviour of the decoder.

To predict the time at which the agent would be at a particular location in Figure 4C (right), we analysed every state s separately, and then averaged over all s. For each s, we identified trials where the RNN passed through s exactly once. We trained a decoder to predict the time at which s was visited and computed the test performance as a function of the true time at which s was visited. As before, all decoders were trained in crossvalidation across current agent location.

Comparisons of RNN performance and efficiency

In Figure 4F-G, we compare three classes of RNNs, which were trained on either the reward landscape task, the moving goal task, or the static goal task with a goal that changed between trials. The performance of each RNN was quantified in each of the three tasks as the probability of choosing the optimal first action (see above). We also computed the average parameter magnitudes of all the networks, . Finally, we computed the average firing rate of each network in the static goal task as . All other RNN analyses apart from Figure S3 used networks trained on the reward landscape task.

Subspace identification

Prior work has investigated the extent to which working memory subspaces are orthogonal (Xie et al., 2022; Dorrell et al., 2024). However, we are primarily interested in the dynamics between subspaces, which are easier to interpret in an orthonormal coordinate system. We therefore asked whether there exists a set of orthogonal subspaces that predict the location of the agent at every time in the future. To do so, we first simulated 6,000,000 trials and collected pairs of neural activity at time t and location at t′(δ) separately for each δ. We did this in two different ways. To estimate ‘planning’ subspaces, subspace δ was defined as a decoder that predicts location at time t′(δ) = δ from neural activity at times t ∈ {−2, −1}. To estimate execution’ subspaces, subspace δ was defined as a decoder that predicts location at t′(δ) = t + δ from neural activity at any t ≥ 0.

We then defined a predictive distribution parametrised by ϕδ for each δ:

Finally, we minimised an objective function that combines the accuracy across δs and the overlap between subspaces:

indicates the normalised parameter vector that predicts being at location si at a delay of δk. The parameters were optimised using ADAM with a learning rate of 5 × 10−3 until convergence or for a maximum of 2000 iterations. We used α1 = 10−4, α2 = 10−3, and αorth was annealed from 0 to 2 × 10−3 over 500 iterations. These hyperparameters were chosen because they resulted in a good approximation to the ‘true’ subspaces in the handcrafted STA.

Estimating effective connectivity

To compute the effective connectivity between representations of different points in spacetime, we projected the learned network parameters into a coordinate system defined by the parameter vectors . Each row of C is a normalised vector that predicts a particular point in spacetime. In this coordinate system, the input weights are given by ; the output weights by ; and the recurrent weights by . To analyse weights between ‘adjacent’ subspaces, we averaged the blocks of that corresponded to weights from any subspace δ to δ + 1 and from any subspace δ to δ − 1. To avoid our analyses being biased by the fact that the networks were trained with supervised learning to predict optimal locations that could only be adjacent, we did not include the weights between subspaces 0 and 1 in this average.

To quantify the similarity between the recurrent weights in this spacetime coordinate system and different order adjacency matrices for the environment, we computed point-biserial correlations. The Δth order adjacency matrix AΔ ∈ ℝN×N was defined as a binary matrix with elements equal to 1 for pairs of states that can be reached from one another in Δ actions, and 0 for all other pairs of states. The 0th order adjacency matrix was defined as the identity matrix.

Perturbation analyses

For the perturbation analyses in Figure 5F-H, we constructed an environment where the reward function R(t, s) was (i) 1.0 for (t, s) ∈ {(0, 0), (1, 1), (2, 2), (3, 6), (4, 10), (5, 10), (6, 10)}, (ii) 0.7 for (t, s) ∈ {(1, 4), (2, 8), (3, 9)}, and -1 for all other points in spacetime. The RNN reliably converged to a representation of the optimal path. We first ran the network dynamics for 10 environment iterations without perturbation after the end of the normal planning period. We then continued to run the network dynamics for 10 environment iterations with a bias term defined by , where α is the stimulation strength. The perturbed representation was defined as the representation at the end of this perturbation period. The two example representations in Figure 5G used α = 0.3 (‘weak’) and α = 10 (‘strong’). The quantification in Figure 5G used a range of α from 0 to 10 for the RNN, and from 0 to 500 for the handcrafted STA. This is because the attractor wells are deeper in the handcrafted STA, and a stronger perturbation is therefore required to push the network out of an attractor state. The control analyses (grey lines in Figure 5G and Figure S5C) were performed by running the same analysis on the RNNs, but with stimulation of the same magnitude in a random direction in neural state space. For the analysis in Figure 5H, we also removed the perturbation and ran the network dynamics for 10 environment iterations after the end of the perturbation period. All analyses in the main text focused on representations in the space of implied future trajectories, . The ‘representational change’ was quantified as the L1 norm of the difference from the spacetime representation at the end of the normal planning period. See Figure S5 for an analysis of the raw firing rates.

RNNs trained in changing environments

For the analyses in Figure 6, we trained another set of RNNs in a version of the reward landscape task where the transition structure changed between trials. The locations of all walls in the environment were provided as a binary input to the agent (see Jensen et al., 2024 for details). For the performance comparison in Figure 6B (top), we evaluated the performance of these networks in the environments that the ‘single structure’ networks had been trained on.

The effective connectivity in Figure 6C-D was computed as in the RNNs trained on a single structure. For these analyses, we identified the subspaces from 1,000,000 trials in a single environment, and repeated this analysis across 30 different environments. We computed similarities between (i) the effective connectivity estimated in an environment and the adjacency matrix of the same environment, and (ii) the effective connectivity and the adjacency matrix from a different control environment. In Figure 6E, the ‘subspace similarity’ was defined as the correlation between the set of vectors that define the subspaces, averaged over all points in spacetime. We computed the similarity between (i) subspaces identified from two sets of independent trials from the same maze, and (ii) the same number of trials from two different mazes. Recall that the effective recurrent weights between subspaces are given by . By using different subspaces in different environments, the RNN changes C between environments, which changes the effective connectivity between pairs of future subspaces (Figure 6E).

For the analyses in Figure 6G-H, we repeated the subspace identification procedure, but with two notable differences. First, we used trials across many environments to find generalised directions in neural state space that predict the future in any environment. Second, we defined the objective function in terms of future transitions instead of locations. Figure 6G quantifies the effective connectivity between future transitions in consecutive subspaces ( and ) that are either ‘consistent’ (k = j), ‘adjacent’ (j ≠k but si and sk are adjacent in an environment with no walls), or any other transitions. In Figure 6H, we first projected the input corresponding to a given wall location wij onto each subspace and normalised the projection within each subspace. We then calculated the dot product between this projection and the normalised directions in neural state space that predict either (i) transitions between si and sj, (ii) transitions from si or sj to some other state sk, or (iii) transitions that do not originate at si or sj. Projection magnitudes were averaged over transitions within each of these three groups, then across subspaces, and finally the mean and standard deviation were computed across 5 RNNs.

eLife Assessment

This important study advances our understanding of the neural substrate of planning trajectories towards a goal by using recurrent neural networks. The manuscript provides solid evidence for most of the claims, but it remains unclear whether the dynamics do indeed bear the defining characteristics of attractors, and the interpretation and scope of some claims may need to be reassessed in light of prior work. The work will be of broad interest to theoretical and systems neuroscientists and to cognitive scientists.

Reviewer #1 (Public review):

Summary:

This work builds a theory to implement planning trajectories towards a goal in a known environment, inspired by analyses of prefrontal neural recordings. Unlike standard neural architectures for this task, such as value-based learning and successor representations, their proposed theory is able to adapt to novel goal locations within a trial. The key to the theory is that future times are represented by orthogonal groups of neurons. The recurrent connectivity between groups of neurons selective to specific future times and locations reflects the learned knowledge of the task. Finally, the authors show that standard networks trained on the task approximate their proposed theory.

Strengths:

The structure of the work is clear, and the presentation of the results is very well written, which is particularly noticeable given the consequential amount of results presented. The authors are able to link their theory with experimental findings in neural recordings. The reverse-engineering of trained recurrent neural networks is very thorough, by analyzing both dynamics and connectivity. The assumptions and predictions of their model are clearly stated.

Weaknesses:

It is unclear whether their proposed theory, "space-time attractors", actually is an attractor network. The authors used recurrent neural networks with very few timesteps, and long single neuron time constants with respect to the task time scales. Attractor networks, as the ones the authors cite, refer to networks that generate nontrivial patterns of activity through recurrent interactions, after long periods of time.

The authors gloss over how the reward inputs are calculated. Computing these reward inputs should be part of the planning process, and the authors are implicitly leaving this problem aside. How does the reward input, which includes future time and location, depend on the actions that have not yet been taken by the agent? It feels like most of the planning computation is already provided by these reward inputs at the beginning of the trial. It could be that the network is only learning to process the planned sequence of actions present in the inputs.

Reviewer #2 (Public review):

This well-written manuscript proposes to use attractors in space and time (STA) as a mechanistic explanation for planning in the prefrontal cortex. The main conceptual hypothesis is that planning is implemented as attractor dynamics in a representation that encodes states at each time step jointly. Depending on inputs, the network relaxes to a trajectory that already contains future states that will be visited at each time step, rather than computing a scalar value at each point in time and space like other classical approaches from RL. The authors compare this approach to implementations such as TD learning and successor representation, and further show that trained recurrent neural networks on specific tasks involving planning develop structured subspaces resembling the ones postulated in STA.

The idea of treating attracting trajectories unfolding in time as the computational substrate for planning is very interesting and potentially important. The explicit construction of a state x time representational space and its implementation via recurrent dynamics are appealing and convincing in the idealized tasks considered. I found the manuscript to be refreshingly explicit regarding several of the assumptions and limitations of the models, for example, the fact that certain advantages can be viewed as properties of the state space itself and not necessarily of a fundamentally new planning mechanism.

Overall, the manuscript presents a cool attractor model that extends in time and explores its performance in a subset of illustrative tasks involving planning. My doubts concern mostly the interpretation and scope of the claims made in the manuscript. Here are a few comments where I detail my questions/concerns:

(1) The authors nicely discuss that much of the difference between STA and classical TD or SR agents is "in some sense a property of the state space rather than the decision making algorithm," and that TD and SR could in principle be implemented in a comparable space x time representation. This is fair, but it also suggests that the central contribution of the manuscript lies primarily in the representational factorization (state x time tiling) and its dynamical implementation via attractors, rather than in a fundamentally new planning algorithm or theory, mechanistic or not. I think theory should be distinguished from mechanism, and it would therefore help the reader to describe the conceptual advancement more as a novel mechanism or implementation than a novel (mechanistic) theory for decision/planning.

(2) Related to my previous point, I think it would be helpful to position STA more explicitly relative to computational/theoretical literature in which attractor networks encode temporally ordered patterns (so effectively including future times). For example, classical extensions of Hopfield networks with asymmetric connectivity implement retrieval of sequences and ordered transitions between patterns (Sompolinsky & Kanter, 1986). More recently, sequential attractors and limit-cycle dynamics have been constructed in structured recurrent networks by the Morrison group (Parmelee et al., 2021). These works do not implement an explicit discretized state x future-time tiling as in STA and do not specifically discuss the usage for planning. However, they do provide concrete precedents for attractor dynamics over temporally structured trajectories in terms of mechanism. It would be useful to discuss this literature and clarify a little what's new mechanistically in the view of the authors.

(3) A central claim of the manuscript is that space-time trajectories are attractors of the STA dynamics. The manuscript does provide empirical evidence consistent with attractor-like behavior. However, it is not explicitly shown whether trajectory representations persist in the absence of sustained external inputs. So it's not clear to me whether the trajectories should be interpreted as intrinsic attractors of the recurrent system, which can be selected by delivering transient inputs, or whether they must be stabilized by a specific continuous external drive. It would be useful if the author could clarify/discuss this point.

As far as I understand it, reward information is provided as input to specific populations encoding future time steps, and that's essential for rapid adaptation without rewiring connectivity. How such future-time-specific reward inputs would be generated and routed to distinct neural populations isn't entirely clear to me. Since this seems to be an essential component of the model, I think it would be important to discuss more deeply the source and plausibility of these reward signals related to different timesteps.

(4) The authors note that vanilla STA scales linearly with planning horizon, and discuss potentially hierarchical extensions for longer horizons. They acknowledge that learning abstractions remains an open challenge, yet the examples of planning in the manuscript are restricted to very short temporal horizons and limited branching complexity. It is not obvious to me in what cases the current implementation and interpretation of STA remains viable (for example, in terms of relaxation iterations) as the horizon and branching factor increase. Relatively simple planning can be managed by simpler, less costly models/algorithms, whereas complex planning is a lot harder to deal with, and it's something that a mechanistic "theory" should address. In the context of the claims of the paper in its present form, I think this is possibly the most important conceptual and practical limitation in the manuscript.

(5) The RNN analyses show that trained networks develop structured subspaces aligned with future time indices and exhibit perturbation behavior consistent with attractor-like dynamics. The manuscript also explicitly notes differences between the trained RNN and the handcrafted STA (e.g., long-range couplings between subspaces and differences in behavior of lower-value trajectories under perturbation), which I much appreciated. My doubt is on the specificity of this result, as trained RNNs on fixed-horizon tasks can develop latent dimensions correlated with temporal progress within a trial or time-to-goal. I think it would help the reader to clarify whether the results demonstrate that STA-like computations emerge in RNNs trained on planning tasks, or that RNNs generally develop some kind of structured spacetime representations when tasks involve future timesteps and some degree of flexibility in the decisions.

A few more minor points, mainly concerning clarity:

(1) The main dynamical equation combines a log-domain recurrent term, a floor operation, and a log-sum-exp normalization step, followed by exponentiation. The intuition/logic behind this specific formulation could be clarified for the reader. For example tt would be helpful to explain why the recurrent input appears inside a log, and also whether/how these operations relate to any multiplicative constraint.

(2) While the computational cost of successor representation in an expanded NT x NT representation is discussed, the corresponding scaling of STA in terms of number of units and connections (as a function, for example, of the planning horizon) isn't clear to me. Perhaps the authors could compare costs more explicitly.

(3) In the RNN analyses, structured subspaces aligned with future time indices are shown. I couldn't find a quantification of how much variance is captured by the subspaces, relative to other latent dimensions. Adding it would help get a feeling for the strength of the alignment.

Author response:

The following is the authors’ response to the current reviews.

We appreciate the additional clarifications suggested by the reviewers, and we will include these in the final Version of Record.

The following is the authors’ response to the original reviews.

Reviewing Editor Comments:

The reviewers are very enthusiastic about this study, but have pointed out a central issue: are "space-time attractors" really attractors?

The reviewers would be willing to increase the assessment of significance if the comments are properly addressed, and in particular, the issue about space-time attractors.

We thank the editors and reviewers for the feedback on our manuscript and have revised the paper to address their questions and concerns. This document includes (i) an overview of the major changes to the paper, and (ii) point-by-point responses to the reviewers. We have also attached a version of the revised paper that highlights substantial changes to the text.

Briefly, the reviewers asked for improved intuition about the STA representation and dynamics, and its relationship to attractor networks. To address their questions, we have restructured the paper. It now starts by introducing the STA, which has a representation and connectivity that are both handcrafted. The revised manuscript characterises the resulting dynamics and fixed points in more detail, both empirically and analytically. We then introduce a new model that directly optimises the fixed points of a neural network to represent an explicit plan of the future. The optimal weights for inferring such representations resemble the STA connectivity empirically. Finally, we analyse our unconstrained recurrent neural network, which learns both optimal representations and connectivity. As also shown in the original paper, this network learns to implement an algorithm that closely resembles an STA. Together, these results show that attractor networks can infer PFC-like representations of the future, and this is an efficient solution to dynamic planning problems known to depend on PFC.

RE1: Expanded theory of STA dynamics

We have now formalised how the STA relates to a formulation of planning as an inference process over future trajectories, which has been previously proposed in cognitive science and reinforcement learning. We show in the revised paper that the STA dynamics resemble an algorithm for approximate inference in the corresponding probabilistic graphical model. This allows us to characterise the fixed points of the algorithm analytically and relate them directly to a well-established cognitive theory of planning. These analyses help bridge the gap between neural implementation and cognitive computation. They shine new light on previous results in the paper while also providing more intuition for the STA dynamics.

We have also included a new model that directly optimises the fixed points of an attractor network to resemble a posterior distribution over future locations from planning-as-inference. This analysis complements the handcrafted STA, where we impose both the representation and connectivity, and the RNN, where both the representation and connectivity are learned. The new model imposes (i) an explicit spacetime representation, and (ii) the multiplicative structure of a message passing algorithm. We then train the weights associated with the forward and backward messages by gradient descent on the KL divergence between (i) the true posterior marginals and (ii) the approximate distribution over future locations implied by the network representation at the fixed point. Supplementary Figure S2 of the revised manuscript shows that the optimal weights reflect the transition structure of the environment, similar to the handcrafted STA model and the task-optimised RNN. This makes the connection between attractor dynamics and planning-as-inference more explicit by showing that the fixed points of an attractor network can be optimised directly for planning.

RE2: Improved characterisation of fixed points

We have clarified how and why the STA is an attractor network. Attractor networks are defined by the existence of stable fixed points. In ring and grid attractors, there is a continuum of such fixed points in the absence of structured inputs (but often with tonic excitation). In contrast, the STA has a discrete set of input-dependent fixed points. We show explicitly in the revised manuscript how these fixed points depend on the reward inputs to the network, and also how they relate to planning-as-inference.

We are not claiming that the STA is exactly equivalent to continuous ring and grid attractors. Instead, we want to convey the intuition that the connectivity of the STA constrains the possible fixed points to be plausible trajectories through space and time. The reward inputs determine which of these possibilities is an actual fixed point in a given planning problem. This is not unlike ring attractors in the presence of strong visual inputs. The connectivity enforces a single bump of activity, and the visual input ‘yokes’ the bump to an appropriate orientation. These similarities and differences are highlighted in the revised paper.

Finally, we have added a new Figure 3 to the main text that characterises the STA fixed points empirically. This figure:

(a) Shows the evolution of the STA dynamics and convergence to different fixed points in different environments (panels A-B).

(b) Shows that the network can converge to different fixed points on different trials in the same environment. This happens when there are multiple equally good paths to a goal (panels B-D).

(c) Shows that other fixed points also exist that correspond to longer trajectories, but the dynamics of the network bias it towards representations of shorter paths. The STA reliably converges to fixed points representing longer trajectories if it is initialised within their basin of attraction (panel F).

Updated main text:

“Unlike ring and grid attractors, the fixed points of the spacetime attractor depend on tonic inputs. However, the connectivity constrains the fixed points to represent continuous trajectories for any combination of inputs. In this section, we show this empirically. Later, we will see that such connectivity is optimal for planning-as-inference.

To compute a plan, it is necessary to know which states will be rewarding in the future. This reward information is provided as an input to the STA and enables fast adaptation without rewiring the synaptic connections. It alters the fixed points of the recurrent dynamics to only include trajectories that are also associated with high cumulative reward (Figure 3; Methods).”

RE3: Ground truth rewards as an input to the network

Both reviewers asked about the external input to the STA that specifies the reward available at different states in the future. In reinforcement learning and cognitive science, ‘planning’ is usually defined as the problem of computing a trajectory that maximises cumulative future reward, given an initial state and a reward function (e.g. Mattar & Lengyel, 2022). This is similar to many real-life situations, where we have a known but distant goal (win a game of chess, finish our paper before a deadline, …). When such a reward function is known, it remains challenging to determine the sequence of actions to get there. This has been the topic of much previous work in neuroscience, including (i) the successor representation, which combines a trial-specific reward function with stable transition statistics; and (ii) different types of sequential search, which use a known reward function to evaluate different possible future trajectories.

To highlight the importance of planning, even when the reward function is known, Figure 4 of the revised manuscript shows that the STA performs better than a greedy baseline that acts according to the immediate reward input instead of planning to maximise cumulative reward. Planning is therefore distinct from learning or inferring a reward function, which is itself a major open question in cognitive science. While undoubtedly interesting, a solution to this problem is beyond the scope of our paper. That is why we decided to simply provide ground truth rewards as an input to the STA. We have clarified the distinction between planning and ‘reward learning’ in the revised paper, and the supplementary material now includes a discussion of where the reward input to the STA could come from.

All performance quantifications in Figure 4 now include an additional ‘greedy’ baseline (grey bars).

This is an agent that acts according to the immediate future reward. The performance improvement of the STA over this baseline highlights the importance of planning to maximise cumulative reward.

Updated main text:

“There are several possible sources of reward input to a spacetime attractor (Supplementary Note). We focus on planning under a known reward function and therefore assume access to ground-truth rewards.”

Reviewer #1 (Public review):

Summary:

This work builds a theory to implement planning trajectories towards a goal in a known environment, inspired by analyses of prefrontal neural recordings. Unlike standard neural architectures for this task, such as value-based learning and successor representations, their proposed theory is able to adapt to novel goal locations within a trial. The key to the theory is that future times are represented by orthogonal groups of neurons. The recurrent connectivity between groups of neurons selective to specific future times and locations reflects the learned knowledge of the task. Finally, the authors show that standard networks trained on the task approximate their proposed theory.

Strengths

The structure of the work is clear, and the presentation of the results is very well written, which is particularly noticeable given the consequential amount of results presented. The authors are able to link their theory with experimental findings in neural recordings. The reverse-engineering of trained recurrent neural networks is very thorough, by analyzing both dynamics and connectivity. The assumptions and predictions of their model are clearly stated.

We appreciate the encouraging comments and hope our revised manuscript addresses the reviewer’s questions.

Weaknesses

(1.1) It is unclear whether their proposed theory, "space-time attractors", actually is an attractor network. The authors used recurrent neural networks with very few timesteps, and long single neuron time constants with respect to the task time scales. Attractor networks, as the ones the authors cite, refer to networks that generate nontrivial patterns of activity through recurrent interactions, after long periods of time.

See RE1 & RE2 for a comprehensive response to this question. Briefly, we show in the revised manuscript how the fixed points of the STA dynamics relate to planning-as-inference, and we clarify the similarities and differences between the STA and other attractor networks in the main text. We show in the new Figure 3 that (i) representations of future paths are stable over long periods of time, and (ii) multiple fixed points can exist when there are multiple paths to the goal. It is also worth noting that the RNN representation in Figure 6H remains stable for 75 time constants and recovers from perturbations. This is substantially longer than during training, where ‘planning’ lasted up to 14 network time constants, and it suggests that the network representation is a stable fixed point.

(1.2) The authors gloss over how the reward inputs are calculated. Computing these reward inputs should be part of the planning process, and the authors are implicitly leaving this problem aside. How does the reward input, which includes future time and location, depend on the actions that have not yet been taken by the agent? It feels like most of the planning computation is already provided by these reward inputs at the beginning of the trial. It could be that the network is only learning to process the planned sequence of actions present in the inputs.

See RE3 for a comprehensive response to this question. Briefly, ‘planning’ is often defined as the problem of computing a trajectory that maximises future reward, given a reward function, initial state, and transition function. The reward function provided to the agent indicates which future states it would be desirable to reach, but not how to reach them. To make this point clearer, we show in Author response image 1 that the representations computed by the STA generate better behaviour than an agent acting greedily according to the reward function specified by the inputs. This highlights the importance of considering distant goals when choosing immediate actions.

Reviewer #1 (Recommendations for the authors):

The text is very nicely written, and I appreciated the way in which methods are presented, with a clear structure and a logical chaining of the different sections. My comments and suggestions refer mostly to the methods and the RNN implementation. Please find below a list of issues.

Relatively major:

(1.3) All the equations of the dynamics should be written in discrete time and not in continuous time. There is no notion of "iteration" in continuous time, so it is currently very hard to understand how the RNN works, and what the different epochs are ("the RNN performed 10 network iterations..."?).

We have rewritten all equations in discrete time and clarified the notion of ‘iterations’.

(1.4) This is pointed out in the public review, but the authors insist on making an analogy between the spacetime attractor implementation of planning and attractor networks. It seems to me that these two types of models are very different. What defines attractor networks (such as grid- or ring-attractor networks) is that recurrent connections internally generate stable states of activity for long periods of time, in the absence of inputs. Nothing like that is shown here. Robust input-driven trajectories are neither necessary nor sufficient for showing that an RNN is an attractor network. The fact that, given the inputs, networks are run for very short periods of time in this work seems to indicate that this is a very different type of network compared to the attractor networks mentioned previously.

See RE1 & RE2 for a comprehensive response to this question. Briefly, it is correct that the fixed points of the STA depend on the inputs, which is different from canonical ring and grid attractors.

We show that the fixed points of the recurrent STA dynamics are reward-maximising paths when conditioned on those inputs. Briefly, we now (i) analytically characterise the input-dependence of the fixed points and show how they relate to planning-as-inference; and (ii) show empirically that STA representations remain stable for long periods of time (Figure 3). The revised paper also clarifies the similarities and differences to previous attractor models.

Minor

(1.5) It would be nice to show more clearly what the inputs are in a given trial, and how they change over time in the RNN (specifying the planning and execution phases). A supplementary figure may help.

How is the information about walls provided to the network exactly? More generally, it would be nice to clearly indicate in the Methods what all the inputs "x" to the RNN are, and how they change over time (during planning, during execution, and how they change as the environment steps are updated).

We have made a new Supplementary Figure S3 that illustrates the inputs to and outputs from the different models. Briefly, the information about the walls is provided to the RNN as a binary vector xw ∈ ℝ 2𝑁. The elements of this vector indicate for each of the N states whether there is a wall (i) to the right of, and (ii) above it. In each trial, a subset of these is present, and a subset is absent. All ‘present’ walls are assigned a value of +1 in xw , and all absent walls are assigned a value of 0 in xw . We have also clarified this in the revised Methods.

(1.6) It would help to clarify, at least in the methods, the shape of all the matrices and vectors that are trained.

We have clarified the shapes of all matrices and vectors in the Methods.

(1.7) N in the methods is not defined (I think N = 16, the total number of locations on the grid).

N is indeed the total number of locations in the state space. This is 16 for almost all analyses in the paper, which involve planning on a 4x4 grid. The updated manuscript includes a few analyses in larger environments, where N is larger. We have clarified this in the Methods.

Reviewer #2 (Public review):

This well-written manuscript proposes to use attractors in space and time (STA) as a mechanistic explanation for planning in the prefrontal cortex. The main conceptual hypothesis is that planning is implemented as attractor dynamics in a representation that encodes states at each time step jointly. Depending on inputs, the network relaxes to a trajectory that already contains future states that will be visited at each time step, rather than computing a scalar value at each point in time and space like other classical approaches from RL. The authors compare this approach to implementations such as TD learning and successor representation, and further show that trained recurrent neural networks on specific tasks involving planning develop structured subspaces resembling the ones postulated in STA.

The idea of treating attracting trajectories unfolding in time as the computational substrate for planning is very interesting and potentially important. The explicit construction of a state x time representational space and its implementation via recurrent dynamics are appealing and convincing in the idealized tasks considered. I found the manuscript to be refreshingly explicit regarding several of the assumptions and limitations of the models, for example, the fact that certain advantages can be viewed as properties of the state space itself and not necessarily of a fundamentally new planning mechanism.

Overall, the manuscript presents a cool attractor model that extends in time and explores its performance in a subset of illustrative tasks involving planning. My doubts concern mostly the interpretation and scope of the claims made in the manuscript. Here are a few comments where I detail my questions/concerns:

We appreciate the enthusiasm about the manuscript and its potential importance. We address the remaining questions and concerns below.

(2.1) The authors nicely discuss that much of the difference between STA and classical TD or SR agents is "in some sense a property of the state space rather than the decision making algorithm," and that TD and SR could in principle be implemented in a comparable space x time representation. This is fair, but it also suggests that the central contribution of the manuscript lies primarily in the representational factorization (state x time tiling) and its dynamical implementation via attractors, rather than in a fundamentally new planning algorithm or theory, mechanistic or not. I think theory should be distinguished from mechanism, and it would therefore help the reader to describe the conceptual advancement more as a novel mechanism or implementation than a novel (mechanistic) theory for decision/planning.

We respectfully disagree that ‘theory’ has to be distinguished from ‘mechanism’. We do agree that ‘computation’ and ‘mechanism’ can often be distinguished. However, we think theories can live at either of these (and other) levels of explanation. What we propose is indeed a potential mechanism for planning that combines recently characterised prefrontal spacetime representations with attractor dynamics to infer desirable ‘plans’. As we show in the revised manuscript, this mechanism resembles the computation of ‘planning-as-inference’, which has previously been proposed in cognitive science (e.g. Botvinick & Toussaint, 2012). Our theory is therefore not about the computation – it is about the mechanism. The title “A mechanistic theory of planning…” is meant to clarify what level of description our paper addresses.

As an example of the importance of mechanistic theories, the computation of angular velocity integration can be implemented in many different ways. Seminal work by Skaggs et al. (1994) and others in the 1990s showed how it can be implemented in neural networks, inspired by experimental data. These theories paved the way for detailed experimental characterisations of the fruit fly head direction circuit more than two decades later (Turner-Evans et al., 2017; Kim et al, 2017; and others). Inspired by this and other success stories, we think an important role of theoretical neuroscience is to develop theories about neural mechanisms that can be tested in future experiments!

(2.2) Related to my previous point, I think it would be helpful to position STA more explicitly relative to computational/theoretical literature in which attractor networks encode temporally ordered patterns (so effectively including future times). For example, classical extensions of Hopfield networks with asymmetric connectivity implement retrieval of sequences and ordered transitions between patterns (Sompolinsky & Kanter, 1986). More recently, sequential attractors and limit-cycle dynamics have been constructed in structured recurrent networks by the Morrison group (Parmelee et al., 2021). These works do not implement an explicit discretized state x future-time tiling as in STA and do not specifically discuss the usage for planning. However, they do provide concrete precedents for attractor dynamics over temporally structured trajectories in terms of mechanism. It would be useful to discuss this literature and clarify a little what's new mechanistically in the view of the authors.

We agree that this is not the first use of attractor networks to represent or compute sequences. Instead, we show that a combination of spacetime representations with attractor dynamics is sufficient to compute plans in dynamic problems known to depend on prefrontal cortex. As the reviewer points out, the primary difference from most previous work lies in the fact that the entire sequence is encoded in a single fixed point of the STA dynamics. This differs from e.g. Sompolinsky & Kanter, where the population encodes one element at a time and generates sequences as limit cycles. The instantaneous encoding of an entire sequence in the STA is what enables planning through parallel message passing rather than sequential search. This is highlighted in the main text of the revised manuscript, which also includes a supplementary discussion of the similarities and differences between the STA and related work on sequences in attractor networks.

Updated main text:

“Entorhinal grid cells are also embed a world model in their connectivity (McNaughton et al., 2006), but they only encode a single location at a time (Vollan et al., 2026). Such networks can generate sequences, but the individual elements are represented one by one (Sompolinsky and Kanter, 1986; Kleinfeld, 1986; Widloski et al., 2025). The spacetime attractor suggests that circuit principles in prefrontal cortex resemble other cortical areas that use structural knowledge to infer features of the world. The major difference is that PFC instantaneously represents many points in time, which generalises known circuit principles to complex planning.”

(2.3) A central claim of the manuscript is that space-time trajectories are attractors of the STA dynamics. The manuscript does provide empirical evidence consistent with attractor-like behavior. However, it is not explicitly shown whether trajectory representations persist in the absence of sustained external inputs. So it's not clear to me whether the trajectories should be interpreted as intrinsic attractors of the recurrent system, which can be selected by delivering transient inputs, or whether they must be stabilized by a specific continuous external drive. It would be useful if the author could clarify/discuss this point.

We show in the revised paper that the fixed points of the STA dynamics take the form r δ = eRδ ◦ (Arδ−1) ◦ (AT rδ+ 1) (Methods). Here, r δ is the activity of neurons representing expected locations in δ actions; R δ is the reward function in δ actions; and A is the environment adjacency matrix. These fixed points depend on the reward inputs through the first term. In the absence of reward inputs, the fixed points are ‘diffusive’, while still respecting the transition structure of the environment. In the presence of reward inputs, they concentrate probability mass on trajectories with high expected reward. We have clarified these properties in the main text and introduced a new Figure 3 that characterises the fixed points of the STA in more detail. See also RE1 and RE2.

(2.4) As far as I understand it, reward information is provided as input to specific populations encoding future time steps, and that's essential for rapid adaptation without rewiring connectivity. How such future-time-specific reward inputs would be generated and routed to distinct neural populations isn't entirely clear to me. Since this seems to be an essential component of the model, I think it would be important to discuss more deeply the source and plausibility of these reward signals related to different timesteps.

See RE3 for a comprehensive response to this question. Briefly, ‘planning’ is often defined as the problem of computing a trajectory that maximises future reward, given a reward function, initial state, and transition function. The reward function provided to the agent indicates which future states it would be desirable to reach, but not how to reach them (see Author response image 1). We agree that the challenge of estimating future reward is an interesting question, but it is beyond the scope of this paper. We have clarified this distinction in the main text and added a supplementary discussion that speculates about where reward information could originate in biological circuits.

(2.5) The authors note that vanilla STA scales linearly with planning horizon, and discuss potentially hierarchical extensions for longer horizons. They acknowledge that learning abstractions remains an open challenge, yet the examples of planning in the manuscript are restricted to very short temporal horizons and limited branching complexity. It is not obvious to me in what cases the current implementation and interpretation of STA remains viable (for example, in terms of relaxation iterations) as the horizon and branching factor increase. Relatively simple planning can be managed by simpler, less costly models/algorithms, whereas complex planning is a lot harder to deal with, and it's something that a mechanistic "theory" should address. In the context of the claims of the paper in its present form, I think this is possibly the most important conceptual and practical limitation in the manuscript.

It is correct that planning gets increasingly challenging with planning depth. In the absence of noise, the STA scales to sequences of up to 12-13 actions – and even longer if the minimum path length is known a priori. Performance gets progressively worse when recurrent activity and parameter noise increase. The revised paper includes a new Supplementary Figure S1A-C that shows how the STA planning ability depends on planning depth for different levels of noise.

We do not consider planning depth to be a major limitation of the work, since humans are rarely thought to plan much more than 6 steps into the future at a single level of abstraction (e.g. van Opheusden et al., 2023). Instead, we believe that hierarchical planning is used to infer trajectories to distant goals (Eckstein & Collins, 2020). To illustrate this point, we have now implemented a proof-of-principle hierarchical STA in Supplementary Figure S1D-E. This simulation shows how an ‘abstract plan’ inferred by one STA can be treated as a goal to infer a more ‘detailed plan’ in a second STA. In principle, this enables the system to compute plans that are arbitrarily long, provided they can be broken down into chunks smaller than the limits imposed by the analyses in Supplementary Figure S1A-C.

Finally, RNNs learn an STA-like algorithm when trained on dynamic planning problems with a planning depth of 6. It is therefore not clear to us whether simpler and less costly algorithms can be easily implemented in the dynamics of recurrent networks.

(2.6) The RNN analyses show that trained networks develop structured subspaces aligned with future time indices and exhibit perturbation behavior consistent with attractor-like dynamics. The manuscript also explicitly notes differences between the trained RNN and the handcrafted STA (e.g., long-range couplings between subspaces and differences in behavior of lower-value trajectories under perturbation), which I much appreciated. My doubt is on the specificity of this result, as trained RNNs on fixed-horizon tasks can develop latent dimensions correlated with temporal progress within a trial or time-to-goal. I think it would help the reader to clarify whether the results demonstrate that STA-like computations emerge in RNNs trained on planning tasks, or that RNNs generally develop some kind of structured spacetime representations when tasks involve future timesteps and some degree of flexibility in the decisions.

An important point to note is that the subspaces we identify do not encode time-to-goal, since they are all active at the very beginning of the trial. We also show that RNNs trained on simpler static tasks do not learn the same algorithm (Supplementary Figure S7) and do not generalise to dynamic problems (Figure 5F). Finally, other algorithms are capable of solving the dynamic problems we study (e.g. the ‘value agent’ in Figure 5B-E). We therefore do not think it is trivial that RNNs learn an STA-like algorithm.

We do think that ‘structured spacetime representations’ generally emerge in RNNs trained on tasks that involve flexible behaviour in changing environments – in some sense that is the claim we are trying to make. It is known that spacetime representations are optimal for structured sequence memory tasks (e.g. Whittington et al., 2025; Dorrell et al., 2026), and we think this is for exactly the same reason. In sequence working memory, the reward function changes in time – for each action, the reward is only non-zero at the corresponding sequence element. However, the adjacency matrix is uniform for sequence memory – any sequence element can follow any other sequence element – so there is no need for planning. We are therefore not claiming that spacetime representations only emerge in the specific planning task we consider here. Instead, we expand the set of problems solvable by such representations to also include adaptive planning known to depend on prefrontal cortex. We have made this more explicit in the revised manuscript.

Updated main text:

“Together, our analyses show that RNNs trained on a dynamic planning task learn to approximate a spacetime attractor. This was also true across variations in model architecture (Methods; Figure S10; Figure S11). These results extend previous findings that explicit spacetime representations are optimal for sequence memory (Supplementary Note; Whittington et al., 2023; Dorrell et al., 2026; Wang et al., 2025). Additionally, RNNs with too few hidden units to learn a spacetime attractor failed to solve the task (Figure S12), suggesting that other solutions are not readily learned by gradient descent.”

A few more minor points, mainly concerning clarity:

(2.7) The main dynamical equation combines a log-domain recurrent term, a floor operation, and a log-sum-exp normalization step, followed by exponentiation. The intuition/logic behind this specific formulation could be clarified for the reader. For example it would be helpful to explain why the recurrent input appears inside a log, and also whether/how these operations relate to any multiplicative constraint.

The specific form of these equations comes from the intuition that the STA approximates planning as an inference process over future trajectories. We have clarified this in the revised manuscript, which explicitly shows how these equations relate to planning-as-inference as formulated previously (e.g. Botvinick & Toussaint, 2012; Levine, 2017).

(2.8) While the computational cost of successor representation in an expanded NT x NT representation is discussed, the corresponding scaling of STA in terms of number of units and connections (as a function, for example, of the planning horizon) isn't clear to me. Perhaps the authors could compare costs more explicitly.

The memory cost of a spacetime-SR would be (NT)^2 and the computational cost (NT)^3 (it is possible that both of these could be reduced by taking advantage of the structured nature of the spacetime successor matrix, but that is beyond the scope of this work). The memory cost of the STA is NT, and the computational cost is (NT)^2 (each iteration of the network dynamics requires the calculation of T matrix-vector products of size NxN, and the number of steps to convergence is approximately linear in T). We have included this comparison in the Supplementary Discussion of the revised paper.

(2.9) In the RNN analyses, structured subspaces aligned with future time indices are shown. I couldn't find a quantification of how much variance is captured by the subspaces, relative to other latent dimensions. Adding it would help get a feeling for the strength of the alignment.

We have added a new Supplementary Figure S6 to the revised manuscript, which quantifies the variance explained by the future-coding subspaces over the course of a trial. The variance explained by the K dimensions encoded by these subspaces is substantially higher than a random baseline, and it approaches the upper bound given by the top K PCs. Interestingly, the future-coding subspaces all explain a lot of variance early in the execution period. During later stages of execution, only the ‘immediate future’ subspaces explain substantial variance. This suggests that the RNN only maintains information in subspaces that represent times before the end of the trial.

References

Botvinick, Matthew, and Marc Toussaint. "Planning as inference." Trends in cognitive sciences 16.10 (2012): 485-488.

Dorrell, William, et al. "An Efficient Computing Theory of Prefrontal Structured Working Memory Representations." bioRxiv (2026): 2026-02.

Eckstein, Maria K., and Anne GE Collins. "Computational evidence for hierarchically structured reinforcement learning in humans." Proceedings of the National Academy of Sciences 117.47 (2020): 29381-29389.

Kim, Sung Soo, et al. "Ring attractor dynamics in the Drosophila central brain." Science 356.6340 (2017): 849-853.

Levine, Sergey. "Reinforcement learning and control as probabilistic inference: Tutorial and review." arXiv preprint arXiv:1805.00909 (2018).

Mattar, Marcelo G., and Máté Lengyel. "Planning in the brain." Neuron 110.6 (2022): 914-934.

Skaggs, William, et al. "A model of the neural basis of the rat's sense of direction." Advances in neural information processing systems 7 (1994).

Turner-Evans, Daniel, et al. "Angular velocity integration in a fly heading circuit." Elife 6 (2017): e23496.

Van Opheusden, Bas, et al. "Expertise increases planning depth in human gameplay." Nature 618.7967 (2023): 1000-1005.

Whittington, James CR, et al. "A tale of two algorithms: Structured slots explain prefrontal sequence memory and are unified with hippocampal cognitive maps." Neuron 113.2 (2025): 321-333.

Version 2

Abstract

Planning is critical for adaptive behaviour in a changing world, because it lets us anticipate the future and adjust our actions accordingly. While prefrontal cortex is crucial for this process, it remains unknown how planning is implemented in neural circuits. Prefrontal representations were recently discovered in simpler sequence memory tasks, where different populations of neurons represent different future time points. We demonstrate that combining such representations with the ubiquitous principle of neural attractor dynamics allows circuits to solve much richer problems including planning. This is achieved by embedding the environment structure directly in synaptic connections to implement an attractor network that infers desirable futures. The resulting ‘spacetime attractor’ excels at planning in challenging tasks known to depend on prefrontal cortex. Recurrent neural networks trained by gradient descent on such tasks learn a solution that precisely recapitulates the spacetime attractor – in representation, in dynamics, and in connectivity. Analyses of networks trained across different environment structures reveal a generalisation mechanism that rapidly reconfigures the world model used for planning, without the need for synaptic plasticity. The spacetime attractor is a testable mechanistic theory of planning. If true, it would provide a path towards detailed mechanistic understanding of how prefrontal cortex structures adaptive behaviour.

Introduction

While different cortical areas support different functions, common computational principles are shared across many areas. For functions as disparate as sensory processing, spatial reasoning, and language comprehension, features of the environment are inferred from partial information. To do so, structural knowledge about the world must be embedded in synaptic connections (Ko et al., 2011; Burak and Fiete, 2009; Iacaruso et al., 2017; Turner-Evans et al., 2020). This constrains neural circuits to represent meaningful interpretations of the environment, and inputs select between these interpretations. It is not known whether similar principles generalise to complex prospective behaviours, such as planning extended action sequences to achieve a distant goal. Recent recordings from prefrontal cortex give an important clue. When mice or monkeys must execute a sequence of actions, neurons represent the entire sequence concurrently (El-Gaby et al., 2023; Xie et al., 2022). Different neuronal populations encode different steps of the future behaviour. If such a representation could be inferred from inputs indicating goals and constraints, the network could plan the future. Excitingly, this solution would use algorithmic principles similar to those known to infer features of the present in other cortical areas.

This paper has four principal aims. (i) To develop a detailed circuit model that infers explicit representations of the future from partial inputs. (ii) To discern the principles of synaptic connectivity that enable such a model to solve complex problems including planning. (iii) To understand when and why this algorithm succeeds while simpler circuit models fail. (iv) To explore how it relates to pre-frontal representations, connectivity, and function. To this end, we first construct a neural network that combines known PFC representations with structured connectivity to plan the future. We then show that the same connectivity emerges when PFC-like representations are imposed, but the weights are optimised for planning. Finally, this solution is learned by unconstrained neural networks, when both their representations and weights are optimised for challenging planning problems.

Sequence representations in PFC

Recent work has uncovered PFC representations underlying sequence working memory (Xie et al., 2022; El-Gaby et al., 2023; Whittington et al., 2023; Botvinick and Plaut, 2006). Separate neural ‘subspaces’, or groups of neurons, represent the expected or desired state of the world at different steps along the sequence of future behaviour. In other words, some neural subspace ‘A’ represents the present, while subspace ‘B’ represents the immediate future, and subspace ‘C’ the more distant future (Figure 1A). Critically, these subspaces are active simultaneously. Together, they instantaneously represent the entire behavioural sequence. Importantly, El-Gaby et al. (2023) showed that the PFC subspaces representing different steps of the future are not independent. Neuronal correlations reflect the structure of the task being performed, even during rest.

Background.

(A) Recent work has identified explicit sequence representations in prefrontal cortex during working memory (Xie et al., 2022; El-Gaby et al., 2023). When an animal has to execute a behavioural sequence (left), individual neurons represent conjunctions of location and sequence element (centre). Separate populations (planes) therefore represent the expected location at different times in the future. The entire sequence is represented concurrently by the simultaneous firing of different neurons that encode the expected location at each time in the future (right). (B) An example V1 cell fires when its receptive field is aligned with an inferred line (blue), but not for a control stimulus with no inference (orange). Such visual inference is mediated by structural priors embedded in the circuit connectivity, where neurons representing consistent visual features excite each other (Iacaruso et al., 2017; Shin et al., 2023). Figure adapted from Lee and Nguyen (2001). (C) Structural knowledge is embedded in the synapses of the head direction system (centre; Turner-Evans et al., 2020), which constrains the network to represent single angles (Kim et al., 2017). Visual and proprioceptive inputs (left) determine which angle should be represented (right). We suggest that a mechanism like (B) and (C) infers the representation in (A). (D) Prefrontal cortex is particularly important in dynamic environments. Stimulus-response associations and repeated choices are robust to prefrontal lesions (i; ii), and acortical mice can solve spatial memory tasks (iii; Zheng et al., 2024). However, PFC is needed for reversal learning (iv; Walton et al., 2010) and when different goals are important, or ‘rewarding’, at different times (v; Shallice and Burgess, 1991). In multiplayer board games (vi), different resources are valuable at different stages, and opponents can dynamically change their strategy.

Planning by inferring the futur

The correlation structure observed in PFC resembles biological attractor circuits that infer the current state of the world, such as the visual input or head direction of an animal (Figure 1B-C; Ko et al., 2011; Chaudhuri et al., 2019). If similar dynamics acted on the future representation in PFC, it would extend its function beyond sequence memory to situations where entire action sequences are inferred. Planning could be solved by inferring sequences of desired actions from goals and constraints in a recurrent network implementation of ‘planning-as-inference’ (Botvinick and Toussaint, 2012; Levine, 2018; George et al., 2021). The resulting algorithm would naturally cope with dynamic environments, because it represents each time in the future separately. This is intriguing because PFC is particularly important for problems that require the correct behaviour to be expressed at an appropriate time (Figure 1D; Shallice and Burgess, 1991; Volle et al., 2011). Planning via attractor dynamics is also consistent with winner-take-all dynamics identified in frontal cortex during non-sequential behaviours (Ruff et al., 2025; Inagaki et al., 2019). Such planning-as-inference differs from most planning algorithms studied in cognitive science and machine learning, which often rely on sequential search (Callaway et al., 2022; Schrit-twieser et al., 2020). Search is easily adapted to new environments, but it is slow at decision time. In familiar environments, where the structure can be embedded in cortical connections, attractor dynamics provide a complementary mechanism for rapid evaluation of many possible futures in parallel.

A mechanistic theory

In this paper, we show that known features of PFC representations and connectivity are sufficient to implement a powerful planning algorithm with minimal additional assumptions. The resulting ‘spacetime attractor’ (STA) instantiates an explicit world model in the synaptic connections between neurons, which allows it to plan by inferring optimal future trajectories. This algorithm resembles other cortical circuits known to infer features of the present, thereby unifying our understanding of PFC with the rest of cortex. The spacetime attractor excels at dynamic problems with changing reward and transition structures, which PFC is critical for and existing mechanistic models cannot solve. RNNs trained on dynamic tasks implement a spacetime attractor in their dynamics, suggesting that it is an efficient solution. Our findings provide a precise mechanistic theory of adaptive behaviour that reconciles prior work on PFC representations with other known cortical computations and deficits from lesions.

Results

An attractor network in space and time

We now introduce the spacetime attractor in more detail. We first review ring and grid attractors, which illustrate how structured connectivity can guide inference from partial information. We then explain how the spacetime attractor uses these principles to infer future behaviour. Neurons are assumed to encode single environment features, but similar ideas apply when each neuron represents a combination of variables (Clark et al., 2025).

Ring and grid attractors

Ring attractors are neural networks that can only encode circular variables (Zhang, 1996; Ben-Yishai et al., 1995). Different neurons have different ‘preferred orientations’, and the network dynamics have a set of fixed points (stable activity patterns) that encode a particular angle. At these fixed points, neurons are only active if their preferred angle is close to the encoded angle (Figure 2A, bottom left). This is achieved by a network where neurons are mutually excitatory if they have similar preferred orientations, and inhibitory otherwise (Figure 2A, top left). The network dynamics infer which angle is most compatible with noisy inputs such as vision and proprioception (Turner-Evans et al., 2017; Skaggs et al., 1994; Kim et al., 2019; Kutschireiter et al., 2023).

The spacetime attractor.

(A) Left; in a one-dimensional ring attractor, neurons representing similar directions excite each other (top, red), and different directions inhibit each other (blue). Connections are shown for the green example cell. Stable network states (fixed points) have neurons at a particular encoded angle being most active (bottom). Right; grid cells (bottom) emerge from a two-dimensional attractor network (top), where cells with similar preferred locations excite each other (red) and intermediate distances inhibit each other (blue). (B) The spacetime attractor encodes a three-dimensional representation, where neurons have both a preferred location (planes) and delay (horizontal axis). This resembles the PFC representation in Figure 1A. Cells that represent adjacent locations in both space and time excite each other (top, red), and different locations at the same time inhibit each other (blue). Given inputs indicating the current location (mouse) and future reward (cheese), the fixed points represent reward-maximising paths through space and time (bottom). While this example has a single static goal, reward inputs can also differ between subspaces if the reward is dynamic. (C) Initial activity is diffuse (left), followed by convergence to a stable representation of a plan (centre). At convergence, a policy is read out from the subspace that represents the next location along the desired trajectory. When the agent moves to the next state, neural activity updates to represent the remaining plan-to-go (right; El-Gaby et al., 2023).

Grid attractors instead encode two-dimensional locations. This is achieved by neurons that are mutually excitatory if they represent nearby locations and inhibitory if they represent more distant locations (Figure 2A, top right; Fuhs and Touretzky, 2006; Burak and Fiete, 2009). The fixed p oints of the network are hexagonal patterns of activity that resemble canonical grid cells (Figure 2A, bottom right; Hafting et al., 2005). Inputs to the network again determine which location is represented in a particular context (Burak and Fiete, 2009).

Spacetime attractors

Inspired by these known biological attractor networks, we propose the existence of a spacetime attractor in prefrontal cortex. The STA is a network that infers explicit representations of the future, and its fixed p oints represent entire paths through space and time (Figure 2B, bottom; Methods). Similar to ring and grid attractors, this is achieved by connecting STA neurons such that consistent states excite each other and inconsistent states inhibit each other. In the space-time attractor, each neuron has both a preferred location and a preferred delay δ, which determines how far into the future its tuning curve is defined. A neuron with δ = 0 fires w hen t he a gent i s currently at the preferred location of that neuron. A neuron with δ = 3 fires w hen the agent e xpects to be at the preferred location after three actions. In such a network, consistent states can form part of a single trajectory, while inconsistent states cannot. The connectivity between neural populations with preferred delays δ and δ + 1 should therefore correspond to the structure – more specifically the adjacency matrix – of the environment (Figure 2B, top). Unlike ring and grid attractors, the fixed points of the spacetime attractor depend on tonic inputs. However, the connectivity constrains the fixed p oints t o r epresent c ontinuous trajectories for any combination of inputs. In this section, we show this empirically. Later, we will see that such connectivity is optimal for planning-as-inference.

Reward inputs select a trajectory

To compute a plan, it is necessary to know which states will be rewarding in the future. This reward information is provided as an input to the STA and enables fast adaptation without rewiring the synaptic connections. It alters the fixed p oints o f the recurrent dynamics to only include trajectories that are also associated with high cumulative reward (Figure 3; Methods). This resembles a previous proposal that prefrontal cortex can compute stable ‘attraction paths’ from the current state to a distant goal (Duncan, 1990). Importantly, the STA can accommodate problems with time-varying reward structures known to engage PFC (Figure 1D; Shallice and Burgess, 1991; Volle et al., 2011; Carlesimo et al., 2014). This is possible because neural populations that represent different times in the future can receive different inputs. Given a restaurant booking in 2 hours and a cinema ticket in 4 hours, the δ = 2 neurons would receive reward input at the restaurant, and the δ = 4 neurons at the cinema. The network dynamics would then infer a future trajectory, where the agent is at each location at the appropriate time. There are several possible sources of reward input to a spacetime attractor in PFC (Supplementary Note). We focus on planning under a known reward function and therefore assume access to ground-truth rewards.

STA dynamics and fixed points.

(A) STA dynamics on a 3 × 3 grid with a single goal (top). The final plane in each row is a max projection that summarises the encoded path. Time is measured in units of the network time constant τ. The converged representation is a stable fixed point that resembles a posterior distribution over trajectories conditioned on reaching the goal (Methods). (B) In an environment with several disjoint paths to the goal, there is initial competition between representations of the alternatives. The factorised representation then drives convergence to a single trajectory (Methods). (C) The STA converges to different fixed points on different trials in the environment from (B). (D) There is not a strong bias towards either of the two fixed points. (E) If the reward input is set to 0 after convergence in (B), the dynamics relax back to a ‘diffusive’ representation around the initial location. This illustrates the input-dependence of the fixed points. (F) In an environment with both a shorter and a longer path to the goal (top), the STA reliably converges to a representation of the shorter path (centre). This is because the shorter path has a higher likelihood and larger basin of attraction in the network dynamics (Methods). If the network is initialised close to a representation of the longer path, it converges to this solution instead (bottom).

To summarise, the spacetime attractor has four main components: (i) different neural populations represent different times in the future; (ii) the neurons are connected according to the structure of the environment; (iii) the current location is an input to the ‘present’ subspace; and (iv) the reward structure is an input to all future subspaces. The resulting fixed points (Figure 2B, bottom) represent trajectories that maximise cumulative reward, and they resemble an approximate posterior over the future in a probabilistic formulation of planning (Methods). The network dynamics infer future locations through a gradual relaxation process, where reward inputs bias the representation in each sub-space towards high-reward locations, while inputs from neighbouring subspaces constrain their representations to be consistent (Figure 2C; Figure 3).

The STA guides behaviour

After computing a plan, it can be used to inform behaviour. In particular, an agent should take an action that leads to the next location on the inferred trajectory (Figure 2C, middle). The entire representation of the future then moves by one ‘action’, so the state that was previously represented by subspace δ is now in subspace δ − 1. The resulting dynamics resemble a ‘conveyor belt’ that allows the STA to execute entire trajectories without recomputing the policy. Such conveyor belt dynamics have been observed experimentally during sequence working memory (El-Gaby et al., 2023). Importantly, the representation remains a fixed point of the spacetime attractor dynamics, because the location and reward inputs are updated when the agent moves and time passes.

World models for planning

Many planning algorithms apply a ‘world model’ sequentially to simulate the consequences of different actions one at a time. The spacetime attractor instead instantiates multiple copies of a forward and inverse world model in the synapses between different neural populations (Methods). Recurrent dynamics can then arbitrate between many possible futures in parallel. Entorhinal grid cells also embed a world model in their connectivity (McNaughton et al., 2006), but they only encode a single location at a time (Vollan et al., 2026). Such networks can generate sequences, but the individual elements are represented one by one (Sompolinsky and Kanter, 1986; Kleinfeld, 1986; Widloski et al., 2025). The spacetime attractor suggests that circuit principles in prefrontal cortex resemble other cortical areas that use structural knowledge to infer features of the world. The major difference is that PFC instantaneously represents many points in time, which generalises known circuit principles to complex planning.

STAs are flexible planners

Prefrontal cortex is particularly important for flexible behaviour in changing environments (Figure 1D; Burgess and Wu, 2013). To understand whether the spacetime attractor is a good model of planning in PFC, we therefore (i) study its behaviour and performance in such dynamic tasks, and (ii) characterise when and how it differs from existing models. We will see that the spacetime attractor is well-suited to dynamic ‘PFC-like’ problems, which other mechanistic models struggle to solve.

From static to dynamic tasks

We designed a set of four tasks that vary in how much the reward changes in space and time (Figure 4A). The tasks are all embedded in Euclidean space for simplicity of exposition, but the underlying principles generalise to other environments with known structure. Task 1 is simple navigation towards a static goal that remains constant across trials. In task 2, the static goal changes between trials. Task 3 is navigation to a goal that also moves within each trial, where each location is only rewarded when the goal is there. Task 4 generalises the idea of time-dependent goals to a non-binary reward landscape. Reward magnitudes are sampled independently between −1 and +1 for each location at each time point in each trial. The objective is to maximise cumulative reward over the trial, which requires balancing immediate reward with the potential for future reward (Figure 4A, right). This is reminiscent of the example from Figure 1D, where the restaurant and cinema are desirable at different t imes. To better understand how the STA solves these tasks, we compare it to two models commonly studied in systems neuroscience. The first is temporal difference (TD) learning, which has neural correlates in striatum (Sutton, 1988; Schultz et al., 1997). The second is the successor representation (SR), which has neural correlates in hippocampus (Dayan, 1993; Stachenfeld et al., 2017). These models assume fixed rewards across trials (TD) or within a trial (SR). Unlike the spacetime attractor, they do not generalise well when the reward changes rapidly.

Model comparisons in static and dynamic tasks.

(A) Example tasks with different degrees of dynamic reward. Colours indicate reward at each location (white to green), and red arrows indicate the optimal policy. In the ‘moving goal’ task, the agent intercepts a target that moves along a different trajectory in each trial (arrow). In the ‘reward landscape’ task, the reward is sampled independently in space and time. The agent has to maximise cumulative reward, and it can be optimal to forgo immediate reward for a later payoff. (B) Representations and performance in the static goal task with a fixed goal. Left: value functions learned by TD and SR agents. Centre: STA representation encoding a path through space and time. Right: performance of each model (see Figure S1 for STA performance in larger environments). The grey bar is a ‘greedy’ baseline that maximises immediate reward, and the horizontal line is a random baseline. (C) Task with a static goal that changes between trials. (D) Moving goal task. Left: the SR computes a value function that averages reward across time-within-trial. Middle: the STA takes into account the moving goal and computes a path that intercepts it. Right: performance of each agent. (E) Performance in the reward landscape task. All error bars indicate 1 standard deviation across 20 agents and environments (dots).

Simpler models can solve static tasks

TD learners gradually propagate value from rewarded locations to all other locations, and optimal policies are only learned when rewards are constant across both time and trials (Figure 4B-C, orange). The SR agent uses the environment structure and trial-specific reward function to compute values, and it can solve tasks where the goal changes between trials (Figure 4B-C, green). The STA can also solve these tasks, but it does so by inferring an entire spatial trajectory to the goal instead of computing a value function (Figure 4B-C, blue). Because the reward is constant throughout a trial, each sub-space receives the same reward input. However, the inferred representation differs across subspaces because it is also constrained to satisfy the current agent location and the transition structure of the world. As we will see, the inference of an explicit spacetime representation by the STA increases its flexibility compared to the amortised state values computed by TD and SR agents.

Only the STA solves dynamic task

While the SR agent can adapt to rewards that change across trials, the reward structure still needs to remain constant for the duration of the planning horizon. This is because the SR computes time-averaged occupancy and lacks fine-grained information about when the agent is where. When the reward changes within a trial, the SR averages values across time-within-trial and fails to intercept the goal at the correct point in space and time (Figure 4D). The STA can solve the moving goal task because the input to each subspace is the reward at that specific moment in time. The representation is therefore biased towards coinciding with the goal in both space and time, and the network dynamics relax to a future trajectory that correctly intercepts the goal (Figure 4D). The advantage of the STA is exacerbated in the more dynamic reward landscape task, where very different locations can be rewarded at different times (Figure 4E). These results suggest that different algorithms could contribute differently to decision making. TD learning could drive rapid decisions in environments with stable rewards; the SR would be more flexible when rewards change on intermediate timescales; and the STA could enable adaptive behaviour in environments with rapid changes. Finally, sequential search would facilitate slower planning in novel environments and could be guided by partial plans from a spacetime attractor.

STAs are efficient planners

We have now seen that spacetime attractors excel at dynamic planning problems. This required (i) a representation of the future, and (ii) structured weights that reflect the environment structure. However, it is not known whether this solution is optimal. In this section, we first train neural networks to infer optimal plans while imposing an explicit representation of the future. These networks learn to embed the environment structure in their recurrent weights. We then train fully unconstrained networks to generate optimal actions, and we show that they learn to infer explicit representations of the future using such structured connectivity.

Optimal weights for future representations

The representation and connectivity of the STA are inspired by theories of planning as an inference over future locations. However, the proposed weights may not be optimal for this computation. We therefore trained a recurrent network to perform inference over representations of the future, which were constrained to be factorised over time. The exact posterior for each future time step has a known functional form, which combines two factors that reflect constraints from earlier and later times in the plan. These factors can be approximated by recurrent inputs from earlier and later subspaces respectively (Derivation). However, we do not know the optimal form of the recurrent weights. We therefore optimised the weights numerically to maximise the similarity between (i) the exact posterior over future locations, and (ii) the fixed points of the network dynamics (Methods). The resulting connectivity resembled the handcrafted spacetime attractor (Figure S2). In particular, neurons representing consecutive time steps were connected according to the environment structure. Unlike the handcrafted STA, more distant populations were also weakly connected. This connectivity captured the higher-order structure of possible trajectories.

Optimal representation and connectivity

So far, we have directly imposed an explicit representation of the future. However, this representation requires many neurons. We investigated whether superior representations exist by training recurrent neural networks (RNNs) to solve dynamic planning problems efficiently (Mante et al., 2013; Stroud et al., 2023). We will see that regularised RNNs learn to solve these ‘PFC-like’ problems with both representations and connectivity that resemble a spacetime attractor, which suggests that it is an efficient solution. The RNNs were trained with supervised learning to generate optimal actions from inputs that indicated the current location and trial-specific rewards (Figure S3; Figure S4; Methods; Supplementary Note). The loss function only constrained the upcoming action and did not impose a representation of the distant future. To avoid biasing the RNN representation, reward information was only provided during initial ‘planning’ prior to the ‘execution phase’ of the task. The objective function also penalised large firing rates and network weights to encourage an efficient solution.

The defining features that allow a spacetime attractor to plan via inference are: (i) it has an explicit representation of the future; (ii) the connectivity embeds the environment structure; and (iii) attractor dynamics constrain the representation. RNNs trained on the reward landscape task exhibit all of these properties. These results are consistent with a prominent theory that PFC resembles a recurrent meta-learner, which learns connections over long timescales that rapidly adapt to new problems through recurrent dynamics (Wang et al., 2018). The STA similarly embeds stable structure in its weights and uses recurrent dynamics for fast planning with different rewards. We show that an STA is the mechanism implemented in the dynamics of meta-learners trained on challenging planning tasks.

RNNs learn explicit future representations

Spacetime attractors compute explicit representations of the future (Figure 5A-B). Critically, these representations generalise over trajectories. In other words, the representation in subspace δ depends only on the expected location in δ actions, and not on the rest of the trajectory (Figure 5A, right). To confirm that the RNN learned such a representation, we trained linear decoders to predict the future from its hidden state. To ensure generalisation across trajectories, we trained the decoders while holding out each ‘current location’ and computed the test accuracy only from those held out locations. The RNN had an explicit representation of the future during both planning and execution (Figure 5C-D; Figure S5). These representations were high-dimensional and explained substantial variance in the RNN activity (Figure S6). Consistent with experimental recordings from PFC (Chiang and Wallis, 2018), the STA and RNN both represented the future more strongly than the past (Figure 5D). In the spacetime attractor, the same subspace always encodes locations δ actions into the future. The future representation of the RNN also exhibited such ‘conveyor belt’ dynamics (Figure 5E). The neural representation of the RNN is thus consistent with a spacetime attractor.

RNNs learn spacetime representations in the reward landscape task.

(A) An STA infers the entire future during planning and represents the ‘future-to-go’ during execution (left). The representation in each subspace generalises across all trajectories passing through that location in spacetime (right). (B) We compare a spacetime attractor; an RNN trained on the reward landscape task; and an agent that computes an exact value function in space and time. The value-based agent computes an optimal policy from ‘neural activity’ containing (i) the value function, (ii) the agent location, and (iii) the time-within-trial (Methods). (C) Decoding accuracy at the end of planning for: left; agent location at each time in the future. Right; the time at which the agent will be at a given location, plotted as a function of the actual time the location was visited. Decoders were trained in crossvalidation across the current agent location (Methods). (D) Decoding accuracy during execution of location at each time in the past or future. (E) We trained a single decoder to predict location at time 3 from neural activity at time 1 (black circle). The same decoder predicted location at time t + 2 from neural activity at any other t, demonstrating ‘conveyor belt’ dynamics. (F) Performance in the static goal (left), moving goal (centre), and reward landscape (right) tasks for RNNs trained on either task (x-labels; colours). (G) Normalised parameter magnitudes of the three RNNs (left) and average firing rates in the static goal task (right). All error bars indicate 1 standard deviation across 5 RNNs (dots).

Intriguingly, RNNs trained on the simpler static and moving goal tasks did not reliably learn explicit future representations (Figure S7). Additionally, the RNN trained on the reward landscape task could solve these simpler tasks, while RNNs trained on the simpler tasks failed in the reward landscape task (Figure 5F). Alternative solutions therefore exist in the simpler tasks, and these solutions are favoured by a pressure to be energetically efficient (Figure 5G). We identify a tradeoff b etween generality and efficiency, where th e PFC-like solution is general but needs more neurons and synapses, while simpler algorithms can solve simpler tasks more efficiently. Un like va lue co ding an d related theories, this suggests a reason why PFC occupies such a large cortical territory.

RNNs learn a world model

In a spacetime attractor, subspaces that represent expected future locations are connected according to the environment structure. Testing this prediction in the RNN requires interpretation of the network weights, which is challenging because the functions of individual neurons are unknown. We therefore project the weights into a coordinate system, where the axes are orthogonal directions in neural state space that predict each point in spacetime (Methods). While this does not perfectly recover the true subspaces in the handcrafted STA, it is a good estimate. The projected RNN weights can therefore be interpreted as interactions between representations of different points in spacetime (Figure 6A).

RNNs learn a world model.

(A) The weights of RNNs trained on the reward landscape task are projected into an orthonormal coordinate system with axes that predict different points in spacetime (top). The projected weights are interactions between points in spacetime (bottom). (B) The average recurrent weights between subspaces separated by a single action resemble the environment adjacency matrix (bottom, ‘RNN’). ‘Empirical STA’ is the same analysis performed on approximate subspaces estimated from neural activity in the handcrafted STA. ‘True STA’ indicates weights between the ground truth future-coding subspaces in the handcrafted STA. Green box in ‘True STA’ indicates weights between (i) the single location denoted by a green circle in ‘Prediction’, and (ii) all locations in the subspace indicated by a green square. (C) Input weights to the ‘current’ (δ = 0; left) and a ‘future’ (δ = 2; right) subspace. Consistent with an STA, the current subspace of the RNN receives location input, and future subspaces receive reward information. (D) Average recurrent weights between an example location (green arrow) and other locations at increasing time differences (Δ; planes). Locations in subspaces Δ apart are more strongly connected if they can be reached in Δ actions. Light green circles indicate these Δth order adjacency matrices (Δ = 0 corresponds to the identity). (E) Correlation between (i) the average connectivity between subspaces separated by Δ actions (lines; legend), and (ii) different order adjacency matrices (x-axis). Shading indicates standard deviation across 5 RNNs. (F) Example environment with a high-value path and a lower-value path. (G) Weak stimulation does not affect the RNN spacetime representation, but strong stimulation switches it to the lower-value path. (H) Magnitude of RNN representational change across time and stimulation strength (Methods).

Remarkably, the recurrent weights between adjacent subspaces closely resembled the adjacency matrix of the environment (Figure 6B; Figure S8; mean ± std correlation of 0.91 ± 0.07 vs. 0.72 ± 0.06 for control environments). The RNN thus learned a world model explicitly in its recurrent weights. Consistent with a spacetime attractor, the δ = 0 subspace received input indicating the current agent location, and subspace δ received input indicating the reward in δ actions (Figure 6C). The RNN also learned weaker connections between more distant subspaces (Figure 6D-E). These connections reflected h igher-order t ransition s tatistics, reminiscent of networks optimised explicitly to infer future representations (Figure S2). In summary, the RNN learned all of the structural components necessary to infer future actions by using a world model to integrate expected reward across time.

RNNs learn attractor dynamics

In a space-time attractor, structured connections give rise to attractor dynamics with fixed points corresponding to explicit representations of desirable futures. To demonstrate similar dynamics in the trained RNN, we perturbed its neural activity at the end of planning. This resembles previous experiments in biological attractors (Inagaki et al., 2019; Kim et al., 2017; Vinograd et al., 2024). An attractor network should be robust to small perturbations, and it should be more sensitive to perturbations towards other plausible representations than perturbations along random directions in neural state space. We analysed the RNN in a setting with two ‘good’ paths that were close in value (Figure 6F). Like the handcrafted STA, the RNN initially converged to a stable representation of the better path, which persisted under weak perturbations (Figure 6G; left). Stronger perturbation of a single location on the alternate path caused a discrete change to a representation of this entire path, including in subspaces not directly perturbed (Figure 6G; right). This new representation remained after removing the perturbation in the handcrafted STA, suggesting that it is also a fixed point of the unperturbed dynamics (Figure S9). In contrast, the RNN representation relaxed back to the original fixed point after removing the perturbation (Figure 6H; Figure S9). The unperturbed RNN dynamics therefore only have a single stable fixed point, suggesting that the network may be less susceptible to local minima.

Together, our analyses show that RNNs trained on a dynamic planning task learn to approximate a spacetime attractor. This was also true across variations in model architecture (Methods; Figure S10; Figure S11). These results extend previous findings that explicit spacetime representations are optimal for sequence memory (Supplementary Note; Whittington et al., 2023; Dorrell et al., 2026; Wang et al., 2025). Additionally, RNNs with too few hidden units to learn a spacetime attractor failed to solve the task (Figure S12), suggesting that other solutions are not readily learned by gradient descent. The spacetime attractor is thus an efficient solution to dynamic planning problems, and it is consistent with a prominent theory that PFC can be understood as a recurrent meta-learner.

STAs can generalise across environments

Hardwiring the transition structure of the environment into the synapses of a spacetime attractor seems prohibitive, because it might prevent the network from generalising across different structures (Figure 7A). Here, we show that is not the case. RNNs trained in environments with changing structure still learn spacetime attractors. The network dynamics are adapted to each environment through input-mediated gating of a generic scaffold. This is a simple mechanism for rapid adaptation that preserves the ability to plan under changing rewards.

Spacetime attractors can adapt to changing structure.

(A) Changing structure requires adaptation. (B) Performance of RNNs trained on the reward landscape task in a single maze or with a different structure on every trial, when evaluated in either a single maze (top) or across changing mazes (bottom). (C) Two example mazes (left) and their corresponding adjacency matrices (centre). The effective connectivity between adjacent subspaces in a single RNN (right) reflects the structure of each maze. (D) Average correlation between subspace connectivity in a maze and the structure of either the same (blue) or a random (grey) maze. (E) The spacetime representation is more similar across distinct trials from the same maze than trials from different mazes (left). By using slightly different subspaces in different environments, a network can match its connectivity to the environment structure (right; orange vs. blue). (F) Putative mechanism for structural generalisation. Instead of representing each future location, neurons encode expected future transitions (black arrows in example state). Each transition τδij (green) excites other transitions that can follow (τδ+1jk) or precede (τδ−1li) it (red arrows). Structural input to PFC inhibits transitions that are not available in a given environment (blue), preventing planning between states that would otherwise be connected (light red arrows). (G) Effective connectivity between directions in neural state space that encode ‘consistent’ consecutive future transitions (τδij and τδ+1jk; green to red), ‘adjacent’ transitions (green to blue), or any other transitions (green to grey). (H) Projection of the input from wall wij onto representations of future transitions through the wall (τij; dark blue), transitions to other adjacent states (τik; light blue), or any other transitions (grey). All error bars indicate 1 standard deviation across 5 RNNs (dots).

Representations adapt to the environment

To study planning in environments with changing structure, we trained RNNs on a variant of the reward landscape task, where the transition structure was different on every trial. The structure was provided to the RNN as a binary input that indicated whether otherwise adjacent states were separated by a wall (Methods). Importantly, the network had to adapt to each new environment through its dynamics while keeping the weights fixed across environments (Wang et al., 2018; Jensen et al., 2024). The generalising RNNs performed nearly as well as networks trained in a single environment (Figure 7B), raising the question of how generalisation is achieved. The neural representation resembled a spacetime attractor in any single environment (Figure S13B). However, the effective connectivity between subspaces changed between environments. Remarkably, the connectivity reflected the transition structure of whichever environment the agent was currently in (Figure 7C-D). This indicates a general solution that allows recurrent networks to use attractor dynamics for planning in any similar environment, without requiring synaptic plasticity.

A model of generalisation

We can understand how the network adapts to the environment at both the ‘population level’ and at a mechanistic level. At a population level, the directions in neural state space that represent each point in spacetime were more similar on different trials from the same environment than across trials from different environments (Figure 7E, left). This change allows the network to align its representation with the appropriate components of the connectivity matrix (Figure 7E, right). How is this achieved mechanistically? We hypothesised that the network explicitly represents future transitions rather than future locations. In other words, there are directions in neural state space that represent ‘in δ actions from now, move from state si to state sj’, which we denote τδij. This generalises the vanilla spacetime attractor, which has a representation of future locations that is invariant to the subsequent state.

While it requires more neurons, explicitly representing future transitions allows the network to use input-mediated inhibition to adaptively ‘turn off’ transitions that are not available in any particular environment. The learned structure of the remaining transitions is then used for planning (Figure 7F). Importantly, the scaffold only needs to include transitions that are possible in at least one environment. This model makes three predictions that we verified: (i) the network explicitly represents future transitions (Figure S13C), (ii) the representations of future transitions are connected by a world model (Figure 7G), and (iii) input indicating a wall between si and sj specifically inhibits representations of future transitions between those states (Figure 7H). These analyses show that the space-time attractor is an efficient planning algorithm in structurally changing environments, and they suggest a general mechanism for adaptive behaviour.

Discussion

We have taken inspiration from known cortical attractor dynamics and prefrontal sequence representations (Figure 1) to develop a new theory of planning. This spacetime attractor model uses different neural subspaces to represent the expected location of an agent at different times in the future. The subspaces are connected according to the structure of the environment, which allows the network to infer optimal future trajectories (Figure 2; Figure 3). The spacetime attractor solves problems with dynamic rewards, which PFC is important for and simpler algorithms struggle with (Figure 4). It solves these problems using internal representations that closely resemble prefrontal representations during sequence working memory, thereby unifying working memory and planning in PFC. RNNs trained to solve dynamic planning tasks learn to implement the spacetime attractor algorithm in their internal dynamics, suggesting that it is an efficient solution (Figure 5; Figure 6). Finally, spacetime attractors can generalise across environments with different transition structures through rapid gating of a general scaffold to reflect each particular environment (Figure 7).

Experimental predictions

The spacetime attractor is inspired by data and makes precise predictions that can be tested in future experiments:

After planning a behavioural sequence, different subspaces of PFC activity should represent distinct steps of the plan.

Optogenetic activation of neurons in a future sub-space should bias representations and behaviour towards the stimulated state at a delay corresponding to the stimulated subspace.

The effective connectivity between subspaces should reflect the structure of the environment. This could be tested in noise correlations across neurons or by explicit holographic stimulation.

Different patterns of neurons should represent the future in environments with different transition structures.

The neurons active in each environment should be connected according to its structure. Inputs that mediate this gating could come through sensory cortex in observable environments or hippocampus when the structure is learned.

We believe that such hypothesis-driven experiments are critical to uncover mechanistic principles in prefrontal cortex. Although the reward landscape task is conceptually simple, RNNs trained on the task learn very high-dimensional representations (Figure S6; Figure S12). The underlying neural mechanism would be difficult to elucidate by studying low-dimensional projections of neural activity. However, the spacetime attractor provides a concise explanation that can be summarised in just a few equations (Methods). This enabled detailed understanding of both the representations, connectivity, and dynamics of the trained RNNs. We hope similar targeted analyses will uncover new mechanistic insights for flexible behaviour in biological circuits.

Decision making across the brain

Our comparisons of different models across different tasks suggest complementary contributions of different brain regions to decision making. Striatum is thought to implement temporal difference learning (Schultz et al., 1997), which facilitates rapid responses in stable environments. Hippocampus embeds the structure of the world, but it is thought to either average representations over future actions in a successor representation (Stachenfeld et al., 2017) or represent one state at a time in a sequence (Wid-loski and Foster, 2022; Whittington et al., 2020; Jensen et al., 2024). This provides an efficient solution to problems with consistent spatial structure and rewards that change on intermediate timescales. A spacetime attractor in frontal cortex could facilitate adaptive behaviour in dynamic environments with a familiar structural scaffold t hat h as been embedded in PFC. Finally, explicit search would facilitate slower planning in novel environments, and it could be guided by partial plans from a spacetime attractor. While spacetime attractors are particularly useful in dynamic environments, they can also solve simpler problems. It is therefore possible that animals have developed a spacetime attractor because they sometimes need it for planning, and then reuse it in simpler tasks. However, it is also plausible that many laboratory behaviours instead engage simpler algorithms in the basal ganglia or hippocampus. If that is the case, richer spacetime problems may be needed to query the representations used for planning in prefrontal cortex.

Interactions with other planning algorithms

Planning via inference in a spacetime attractor differs from most models of planning in cognitive science. In particular, the spacetime attractor posits that planning can happen via recognition, where the sequence of steps needed to reach some desired state is directly inferred. In contrast, many studies of human planning focus on explicit search, where different paths are simulated and evaluated sequentially. We suggest that these two processes coexist, with spacetime attractor dynamics facilitating rapid planning-as-inference in familiar environments, where the structure has been embedded in prefrontal connections. The STA could also help focus explicit search towards putative high-value paths and evaluate the utility of different paths. This could happen through interactions with hippocampal activity sequences that have been proposed to facilitate planning (either replays or theta sequences; Foster, 2017; Jensen et al., 2024; Vollan et al., 2026). In particular, Jensen et al. (2024) suggest that PFC biases which sequences are generated by hippocampus. PFC would then update its representation to make high-reward sequences more likely in an iterative policy improvement process. Hippocampal replay has also been proposed to implement a ‘DYNA’ algorithm, where past experience is used to learn a model-free policy during rest or sleep (Mattar and Daw, 2018, 2026). This is complementary to both the spacetime attractor and explicit search. DYNA facilitates rapid decision making when the optimal policy remains stable across time, while decision-time planning allows adaptation to changing environments.

Learning a spacetime attractor

In the hand-crafted spacetime attractor, we embedded copies of the environment adjacency matrix in the connections between every pair of consecutive subspaces. The RNN analyses show that such structure can be learned from repeated experience. However, learning copies of the same parameters independently for each pair of subspaces is inefficient. It would be more efficient to store a cache of experienced trajectories that can be used to learn all of the parameters. Hippocampal replay has been proposed to build cognitive maps via such interactions with prefrontal cortex (Bakermans et al., 2023; Ou et al., 2025). In particular, Ou et al. (2025) suggest that hippocampal replay consolidates structural information from hippocampus into cortex. Replay from hippocampus to PFC during sleep or rest could therefore provide a mechanism for learning all of the STA parameters from the same data.

The state space of planning

We have assumed that planning happens at the level of individual locations in the environment. However, humans often plan in more abstract spaces, which improves efficiency by reducing the required planning depth. We also plan hierarchically by first computing an abstract plan that can be refined in increasing levels of detail (Eckstein and Collins, 2020). Multiple spacetime attractors operating at different levels of abstraction can be coupled to implement such a hierarchical planner (Figure S1D-E). One ‘abstract’ STA computes a high-level plan. A second STA treats the next abstract state as a goal and computes a detailed plan to get there. Stacking spacetime attractors in this way allows inference of plans that are exponentially long in the depth of the hierarchy, and therefore the number of neurons. This greatly improves efficiency over the linear scaling of a non-hierarchical STA.

Environment interactions

We have focused on environments that evolve independently of the agent. The rewards and transition structure can therefore be estimated independently and provided as inputs to a spacetime attractor in PFC. Many problems involve environments that instead depend on the behaviour of an agent. In ‘key-door’ problems, moving through a locked door becomes possible only after picking up the key. We posit that such problems can be solved by structural gating of a learned scaffold, similar to the STA that generalises across environment structures (Figure 7). The structural gating would no longer be an external input, but instead a function of the spacetime representation itself. For example, activating a cell that represents ‘being at the key’ could disinhibit ‘transitioning through the door’ at a later time.

Another important example of environment interactions is social behaviours, where agents can change their plans in response to each other. A system of two spacetime attractors could simulate the joint dynamics of two such agents. The first STA ‘A’ infers future behaviour for agent A. The reward input would be the output of a different STA ‘B’ that predicts the behaviour of another agent. The reward for STA B would itself be an output of STA A, which couples the predicted behaviour of the two agents. The combined system relaxes to a fixed point, where the behaviour of A is optimal given the predicted behaviour of B and vice versa – a putative neural implementation of ‘theory of mind’.

Outlook

We have developed a new theoretical framework for planning in prefrontal cortex. It builds on recently discovered prefrontal working memory representations and known attractor dynamics in other neural circuits. The spacetime attractor extends these principles to prospective behaviours in complex and changing environments – a setting that has previously eluded mechanistic circuit models. Existing data does not allow us to test these ideas explicitly. Instead, we have provided a series of concrete predictions for future experiments. We hope this will inspire new work in both experimental and computational neuroscience.

Methods

The spacetime attractor

Problem specification

We focus on the problem of planning, which can be defined as computing a sequence of actions that maximises expected future reward. To solve this problem, it is necessary to know (i) the current state st of the agent; (ii) the transition structure of the environment; and (iii) which states are likely to be rewarding in the future. We assume that the transition structure remains constant throughout each trial and use A ∈ ℝN×N to denote the adjacency matrix of the environment. Here, N is the number of states, and A has elements Aij = 1 if it is possible to move from state st = j to st+1 = i, and Aij = 0 otherwise. We allow the reward function R ∈ ℝT ×N to vary over the course of a trial, such that Rδi indicates the reward associated with visiting state st+δ = i in δ actions, and T is the fixed length of the planning horizon. The challenge of estimating future rewards from past data is beyond the scope of this paper, and we simply assume access to a ground-truth reward function. Instead, we show how a recurrent neural network can embed A in its weights and take st and R as inputs to compute a trajectory {st, st+1, …, st+T} with high cumulative reward ∑δ=0TRδ(st+δ).

Network dynamics

The spacetime attractor is a recurrent neural network with an exponential nonlinearity and block-wise normalisation. We define zδi as the ‘potential’ and rδi as the ‘firing rate’ of a neuron that represents a plan to ‘be at location st+δ = i in δ actions’. Each iteration of network dynamics then involves three steps:

τΔzδi=−zδi+βRδi+log⁡(∑jAijrδ−1,j)+log⁡(∑kAkirδ+1,k) Update

zδi=zδi−log⁡∑jezδj Normalisation

rδi=ezδi Nonlinearity

Here, τ = 50 is the time constant of the dynamics, Aij are weights corresponding to the adjacency matrix, and Rδi is an input that reflects the reward available at location st+δ = i in δ actions. To anchor the plan at the current agent location, we set R0i = 20 if st = i and 0 otherwise. β is a temperature parameter that determines how strongly the agent values reward, and we set it to β = 9 unless otherwise noted. We take the policy of the agent to be greedy in the representation in subspace δ = 1 of locations accessible from the current state: a = argmaxi∈ 𝒩 (s) [r1i]. The action a specifies which location to move to next. We used a maximum planning horizon of T = 6 for all analyses in the main text, a planning horizon of T = 16 in Figure S1B, and a planning horizon of T = 32 in Figure S1C. The normalisation in line (2) can be implemented through a subspace-specific inhibitory neuron with a logarithmic nonlinearity (Iδ=log⁡[zIδ]), which receives input from all excitatory neurons in the subspace (z˙Iδ=−zIδ+∑irδi) and provides reciprocal inhibition. In practice, we normalised analytically for simplicity of implementation.

The fixed points of the STA dynamics have the form (1) rδi∗∝eβRδi(∑jAijrδ−1,j∗)(∑kAkirδ+1,k∗). At convergence, the representation is therefore biased towards high-reward states (first term), while also being constrained to represent continuous trajectories (second and third terms).

Relationship to planning-as-inference

Here we briefly outline how the STA dynamics are inspired by previous models of ‘planning-as-inference’. Following Botvinick and Toussaint (2012), we formulate planning as an inference problem in a Markovian probabilistic graphical model. We assume a uniform prior over possible actions, such that p(st+1=i∣st=j):=Tij=Aij∑kAkj. In line with Solway and Botvinick (2012) and Levine (2018), we define observations o ∈ [0, 1] as binary ‘optimality variables’ with emission probabilities (2) p(ot+δ=1∣st+δ)∝eβRδst+δ (3) p(O∗∣{st+δ}δ=0T)∝e∑δ=0TRδst+δ, where we define O∗:=⋂δ=0T[ot+δ=1] as the condition that ot+δ = 1 for all δ. States with high reward are therefore more likely to have o = 1 than states with low reward, and the probability that all observations take a value of 1 is monotonic in the reward associated with the entire future trajectory. We can now condition on 𝒪 ∗ to compute a posterior distribution over future states: (4) p({st+δ})∣O∗)∝p(O∗∣{st+δ})p({st+δ}). The posterior marginals can be computed with the forward-backward algorithm, but that requires three separate neural populations for each time in the plan to represent forward messages, backward messages, and their product. Instead, we define a factorised distribution Q(st+δ) = Π qδ(st+δ) and optimise it to resemble the true posterior as closely as possible. We assume that each component of this distribution is encoded in the firing rates rδ ∈ ℝN of a single population of neurons, such that qδ(st+δ = i) := rδi.

It is common to factorise the true posterior distribution over trajectories as: p({st+δ}∣O∗)∝p(st)p(ot=1∣st)(∏δ=1Tp(ot+δ=1∣st+δ)p(st+δ∣st+δ−1)) However, we can instead ‘split’ the transition terms across consecutive factors in a symmetrised factorisation: p({st+δ}∣O∗)∝p(st)p(ot=1∣st)p(st+1∣st)12(∏δ=1T−1p(ot+δ=1∣st+δ)p(st+δ∣st+δ−1)12p(st+δ+1∣st+δ)12)p(ot+T=1∣st+T)p(st+T∣st+T−1)12 We can therefore define a ‘factor’ fδ(st+δ,st+δ−1,st+δ+1):=p(ot+δ=1∣st+δ)p(st+δ∣st+δ−1)12p(st+δ+1∣st+δ)12 for each time in the plan. While the approximate qδ only depends on st+δ, fδ also depends on st+δ− 1 and st+δ+1. We want our approximate posterior to resemble the product of true factors as closely as possible: (5) ∏δ=0Tqδ(st+δ)≈∏δ=0Tfδ(st+δ,st+δ−1,st+δ+1). Loosely inspired by expectation propagation Minka (2013), we choose to locally minimise the discrepancy between the true and approximate posteriors. In other words, we define an objective function for component δ as the KL divergence between (i) the factor fδ(st+δ, st+δ−1, st+δ+1), while approximating the distributions over st+δ′ ≠ δ by qδ′ (st+δ′); and (ii) the distribution Q({st+δ}) implied by the firing rates {rδ}. We also introduce a normalisation constant Zδ to the first distribution. This gives rise to a loss function of the form Lδ=KL[Zδ−1fδ∏δ′≠δqδ′‖qδ∏δ′≠δqδ′]=EZδ−1fδqδ−1qδ+1[log⁡fδ−log⁡Zδ+log⁡qδ+1+log⁡qδ−1−log⁡qδ−log⁡qδ+1−log⁡qδ−1]=EZδ−1fδqδ−1qδ+1[log⁡p(ot+δ=1∣st+δ)+12log⁡p(st+δ∣st+δ−1)+12log⁡p(st+δ+1∣st+δ)−log⁡rδ−log⁡Zδ]=Zδ−1∑ijkrδ−1,jTij12rδ+1,kTki12eβRδi(βRδi+12log⁡Tij+12log⁡Tki−log⁡rδi−log⁡Zδ). Here, Tij12 indicates an element-wise square root of each term in the transition matrix. We introduce a Lagrange multiplier λ for normalisation and take derivatives to minimise this objective: (6) ∂(Lδ−λ(1−∑jrδj))∂rδi=λ−eβRδirδiZδ(∑jTij12rδ−1,j)(∑kTki12rδ+1,k). This local optimisation of a factorised approximate posterior leads to the ‘mode-seeking’ dynamics seen in Figure 3B. The fixed points have the form (7) rδi∗∝eβRδi(∑jTij12rδ−1,j∗)(∑kTki12rδ+1,k∗). We now recall the STA fixed points: (8) rδi∗∝eβRδi(∑jAijrδ−1,j∗)(∑kAkirδ+1,k∗). The difference between the two is that the STA uses the adjacency matrix A in place of the element-wise square root of the transition matrix, T12. The STA therefore has the form of a dynamical system with fixed points that locally approximate the posterior marginals over trajectories, conditioned on ‘optimality’ in the planning-as-inference formalism. It also has very similar dynamics to another neural implementation of planning-as-inference by Donnarumma et al. (2025). However, the local approximation with marginals may be quite poor as it neglects parts of the transition matrix. For this reason, Figure S2 and Derivation directly optimise the dynamics of a recurrent network for inference.

Practical considerations

Here we note a few deviations from a simple interpretation of the STA dynamics as approximate inference. First, we use the adjacency matrix instead of the transition matrix or its square root, which differs from a probabilistic interpretation. We do this because the true posterior distribution favours trajectories that pass through low-degree states, which have higher probability under the prior. In contrast, the recurrent weights between RNN subspaces representing consecutive future actions in Figure 6 were generally more similar to the adjacency matrix than the transition matrix. One way to interpret the adjacency matrix is as an action-optimised transition matrix, Aij = maxa p(st+1 = i | st = j, at = a). This generalises the idea to stochastic environments, but it does not optimise a well-defined objective to the best of our knowledge. All results in the paper hold qualitatively when using Tij12 in place of Aij for an implementation that is more faithful to planning-as-inference.

Second, we add a small amount of noise ηij ∼ 𝒰 (− 0.025, − 0.015) to each element of the weight matrix instead of using the exact adjacency matrix. Noise is added because it is unreasonable to expect biological systems to exactly recapitulate the transition structure of the environment. It is restricted to be negative because the STA can be susceptible to discontinuous plans if noisy weights imply a finite probability of transitioning between two distant states (Supplementary Note). We similarly added white noise ξ ∼ 𝒩 (0, σ = 0.1) to the potential of all neurons at each iteration of the network dynamics.

Third, we threshold both the neural potentials (zδi), the forward messages (log Σ j Aij rδ−1,j), and the backward messages (log Σk Aki rδ+1,k) at a minimum value of ϵ = −100 for numerical stability.

Finally, the STA representation updates automatically when the agent moves, since both the location and reward inputs change. However, an additional ‘feedforward component’ in the form of the identity matrix was added to the connectivity between subspaces δ and δ − 1 during the first 100 network iterations (2 time constants) after every action. This is inspired by the ‘update neurons’ of the fruit fly head direction circuit (Turner-Evans et al., 2017) and stabilises the ‘conveyor belt’ dynamics, but it is not necessary for any of the main results in the paper (Supplementary Note). We ran the STA dynamics for 400 iterations before each environment update to ensure convergence to a stable representation.

Optimised approximate planning-as-inference

In the handcrafted STA, we imposed both a representation and connectivity inspired by planning-as-inference. However, our derivation of approximate inference absorbed some of the transition terms into the approximate marginals. The handcrafted connectivity may therefore not be optimal for inference of future representations. Here we describe our approach to learning the weights of a recurrent network that optimally approximates planning-as-inference with a factorised distribution (Figure S2). The network dynamics are given by

τΔz=−z+βR+log⁡(Gr→+b→)+log⁡(Gr←+b←) Update

zδi=zδi−log⁡∑jezδj Normalisation

rδi=ezδi Nonlinearity,

where b→=[1,0,…,0]&b←=[0,0,…1] represent our uniform boundary conditions. Here we use R, r ∈ ℝNT ×1to denote the vectorised versions of the reward function and neural activity across all times and locations. G→,G←∈RNT×NT are constrained to be lower and upper block triangular respectively, and the STA dynamics are recovered if we set their off-diagonal b locks to the adjacency matrix.

The fixed points of these dynamics are given by (9) r∗∝eβR∘(G→r∗+b→)∘(G←r∗+b←). We optimised these fixed points to minimise the KL divergence between (i) the true marginal posterior pδi = p(st+δ = i |𝒪∗) at each delay δ, and (ii) the approximate posterior q(st+δ=i)=rδi∗ implied by the firing rates r∗. The gradient of this objective with respect to the parameters is given by (10) Ω:=diag(eβR). (11) u→:=Gr∗→+b→, (12) u←:=G←r∗+b←, (13) Pt:=1Zt(I−rt∗1⊤). (14) P:=blockdiag(P1,…,PT) (15) J=I−PΩ[diag(u←)G→+diag(u→)G←]. (16) ∂L∂G→=[Ωdiag(u←)P⊤J−T(pr∗)](r∗)⊤. is given by the symmetric expression.

For each parameter update, we ran a batch of 200 trials from the reward landscape task with τ = 10 for a maximum of 500 network iterations or until convergence (maxbatch,δ,i[abs(qbδi′−qbδi)]<10−4). We then calculated the average gradient for G→ and G← across the batch and used ADAM to optimise the parameters. After each parameter update, we projected G→ and G← back onto the set of positive lower and upper block triangular matrices.

Tasks

We used three different classes of tasks to train and compare models in this work – the ‘static goal’ task, the ‘moving goal’ task, and the ‘reward landscape’ task. All tasks required agents to navigate mazes on a 4 × 4 grid with N = 16 states (except the analyses of larger environments in Figure S1). Walls prevented transitions between some pairs of otherwise adjacent states, and the wall configurations were sampled as described by Jensen et al. (2024). For most analyses, the wall configuration remained fixed across all trials, and only the reward function changed. For the analyses in Figure 7, the wall configuration changed across trials, and the agent had to adapt to the new transition structure through its recurrent network dynamics. In all tasks, the agent location was sampled randomly at the beginning of each trial. At the beginning of each trial, the agent location and reward function were ‘frozen’ for 5-7 (randomly sampled on each trial) iterations of the environment. This constituted an initial ‘planning’ phase that was followed by an ‘execution’ phase, where (i) the agent took actions that changed its location, and (ii) the reward function updated as described in more detail below. For all analyses, the full reward function was provided to the agent during the initial planning period. For the handcrafted STA and the RNN in Figure S10, the input provided reward information that was relative in time during execution – it indicated what the reward would be at a given location in δ steps rather than at time t. The reward input was zero for all δ that corresponded to time points beyond the end of the trial. For all other RNN analyses, the reward input was set to zero during execution (Figure S3).

Moving goal task

In this task, a full goal ‘trajectory’ was sampled on each trial. The goal trajectory was sampled as a restricted random walk that could only turn around if it reached a dead end. The start location of the agent was restricted to not coincide with the start location of the goal. The trial terminated when the agent was at the same location as the goal at a given moment in time, or after a maximum of T = 6 actions. The reward input to the agent, R ∈ ℝT ×N, was a matrix indicating the location of the goal at every point in the future. The reward input consisted of ‘+0.6’ for the goal location and ‘-0.6’ for all other locations at each moment in time. In other words, Rδi was ‘+0.6’ if the goal would be at location st+δ = i after δ actions, and −0.6 otherwise.

Static goal task

This task was identical to the moving goal task, except that the goal remained stationary for the duration of each trial. The static goal task was implemented in two different variants – one where the goal remained fixed across all trials, and one where the goal was resampled at random in every new trial. The handcrafted models were evaluated in both of these tasks, while the RNNs were only trained with a goal that changed between trials.

Reward landscape task

In this task, every element Rti of the reward function R ∈ ℝT ×N was sampled uniformly and independently between −1 and +1 at the beginning of each trial. The complete future reward structure was provided as an input to the agent as described above. Every trial finished when the agent had taken T = 6 actions.

Quantification of performance

For most performance comparisons, we computed the probability of choosing the optimal first action in each trial. We define the optimal first action as the first action of the trajectory that maximises cumulative reward over the entire trial. We use this metric rather than the cumulative reward for two reasons. (i) We are interested in the process of planning, whereby an agent balances immediate and long-term reward. This is most challenging for early actions, while the greedy policy is optimal for the last action. (ii) The probability of choosing an optimal action is readily interpretable as a number between 0 and 1. All results were qualitatively similar if we instead used the probability of choosing an optimal action averaged over the entire trial as a performance metric, or if we used the average empirical reward.

For the analyses in Figure S1, we analysed whether the entire trajectory implied by the STA representation was correct. A representation was considered correct if the location with the highest firing rate in each subspace (i) was adjacent to the previous location, (ii) was closer to the goal than the previous location, and had a firing rate of at least 0.3.

Reference models

Here we provide details of the temporal difference and successor representation baselines that we compare the spacetime attractor to. For further details, we refer to Jensen (2023). Common to both of these frameworks is that they explicitly estimate the ‘value function’ under some policy π: (17) Vπ(s)=Eτ∼pπ(τ)[∑t′≥tγt′−trt′∣st=s]. Here, Eτ∼pπ(τ)[⋅] indicates an expectation taken over trajectories τ resulting from the agent following π. The ‘TD’ and ‘SR’ heatmaps in Figure 4 show the computed value functions. In both cases, we take the policy of the agent to be greedy in the value function evaluated at all locations accessible from the current state.

Temporal difference learning

We implement vanilla temporal difference learning, which computes a value function by iteratively applying the update (18) ΔV(st)=α(−V(st)+rt+γV(st+1)). We set the temporal discount factor to γ = 1, since we are interested in maximising the non-discounted cumulative reward. We allowed the TD agent to interact with the environment for 4,000 trials with a learning rate of α = 0.05 before analysing its performance.

Successor representation

In the successor representation, the value function is decomposed as (19) Vπ(s)=Eπ[∑t=0γtrt∣s0=s] (20) =∑t=0γt∑s′pπ(st=s′∣s0=s)r(s′) (21) =rTmsπ. Here, r is a vector of the average reward associated with each state, and msπ is a vector of the expected discounted future occupancy of state s′ if the agent starts in state s and follows policy π: (22) Mss′π=∑t=0γtpπ(st=s′∣s0=s). The full matrix M π, constructed from stacking the msπ corresponding to all states s, is denoted the ‘successor matrix’, and it allows us to write down a vector of expected rewards from all states as (23) vπ=Mπr. M π can be learned using a TD-like algorithm as above. However, since the STA algorithm also has explicit access to the transition structure of the environment, we simply computed the exact successor matrix as the geometric series Mπ = I + γT π + γ2(T π)2 + … = (I − γT π)−1 with γ = 0.95. When computing the successor matrix, we take π to be the diffusion policy, and T π is therefore the diffusion matrix. This is similar to how the adjacency matrix serves as a global structural prior in the STA.

Spacetime value agent

In Figure 5, we compare representations to an agent that computes a value function in a state space consisting of space and time. This agent uses dynamic programming to compute the value of being at location s at time t under an optimal policy for every combination of s and t: (24) V(T,s)=RTs (25) V(t,s)=Rts+maxs′∈N(s)V(t+1,s′). At time t, the optimal policy is greedy in the value of accessible locations at time t + 1. For the decoding analyses in Figure 5, we took the ‘neural representation’ to be the concatenation of (i) the flattened value function vf ∈ ℝTN, (ii) a one-hot representation of the current location, and (iii) a one-hot representation of the current time-within-trial. These three quantities are sufficient to compute an optimal policy.

Recurrent neural networks

In this section, we explain how recurrent neural networks were trained and analysed. All networks were fully connected with Nrec = 800 hidden units (except Figure S12) and ReLU nonlinearities. Each iteration of network dynamics was given by (26) τΔz=−z+Winx+Wrecr+brec+η, (27) r=[z]+ (28) y=Woutr+bout. Here, z∈RNrec are the network potentials, r∈RNrec are the firing rates, x∈RNin are the inputs, θ={Win∈RNrec×Nin,Wrec∈RNrec×Nrec,Wout∈RNout×Nrec,brec∈RNrecc,z0∈RNrec,bout∈Rnout} are the network parameters, and η∼N(0,Σ=10−6INrec) is Gaussian noise. All simulations used a time constant of τ = 5. The output policy π∝ey∈RNout was defined in global allocentric coordinates as the desired next state (Nout = N). In Figure S11, we also analyse a network that produced a policy in ‘local’ coordinates that indicated the desired movement direction (Nout = 5). The input consisted of (i) a one-hot representation xloc ∈ ℝN of the current location in the environment; (ii) the reward xrew ∈ ℝNT as described for the tasks above; and (iii) a binary representation xw ∈ ℝ2N of the location of all walls (Jensen et al., 2024). The elements of xw indicated for each of the N locations whether there was a wall (i) to the right of, and (ii) above it. In each trial, a subset of these were present, and a subset were absent. All ‘present’ walls were assigned a value of +1 in xw, and all absent walls were assigned a value of 0. The total input dimensionality is therefore Nin = N (1 + 2 + T). Gaussian noise with a standard deviation of σ = 10−3 was added to the inputs before passing them to the RNN. A schematic illustration of all RNN inputs and outputs is provided in Figure S3.

For all analyses in the main text, the RNN performed 10 network iterations per environment iteration (action). Reward input was only provided during the planning phase and set to 0 during the execution phase (Figure S3). This was done to avoid biasing the execution-time representation towards a ‘relative’ subspace representation by providing reward input in that format (Supplementary Note). For comparison with this ‘working memory’ RNN, we also trained an RNN with reward input during both the planning and execution phases, which learned qualitatively similar representations and dynamics (Figure S10). For this network, the reward input was given relative to the current time as in the handcrafted STA, and the number of network iterations was randomly sampled between 9 and 11 before each environment iteration.

We optimised all network parameters θ to minimise the loss function (29) L(θ)=λθLparams(θ)+Eτ∼πθ,opt[∑tLacct+λrLratet+λeLentt] (30) Lparams(θ)=|θ|22 (31) Lacct=∑a∈Aoptt−πt(a) (32) Lratet=|rt|22 (33) Lentt=∑aπt(a)log⁡πt(a). Aoptt is the set of optimal actions at time t, and Eτ∼πθ,opt[⋅] indicates an expectation over trajectories induced by the policy of the agent, renormalised over optimal actions. That is, we consider an imitation learning setting where ties between equally optimal actions are broken according to the actual policy of the agent. We used λθ = 2 × 10−7, λr = 10−5, and λe = 10−4 for all analyses. The RNNs were trained using Adam (Kingma and Ba, 2015) with a learning rate of 3 × 10−4 for 200,000 batches of 200 trials (250,000 batches for the RNNs trained in changing mazes). All results are reported as mean and standard deviation across 5 separate RNNs that were trained from different random seeds and with different environment transition structures.

Analyses

In this section, we describe the analyses used to compare different handcrafted models and trained RNNs.

Fixed point analyses

For each analysis in Figure 3, we ran the STA for 1200 network iterations (25 time constants) with recurrent noise ξ ∼ 𝒩 (0, σ = 2.0) and no structural noise added to the weights. To quantify the frequency of convergence to different fixed points, we repeated each simulation 1000 times from a uniform initial representation. A simulation was considered to converge to a particular reference representation if the appropriate location in each subspace had a final firing rate of at least 0.9. For the analysis in Figure 3F (bottom), we set the initial neural potential in subspace δ = 2 to z2i = 0 for the bottom right location and z2i = − 100 for all other locations. The remaining subspaces were initialised to uniform representations. For the simulation in Figure 3E, we initialised the neural potentials to the final state from Figure 3B, set all reward inputs to zero, and ran the network dynamics for 1200 iterations.

Decoding analyses

To decode future locations from neural activity (Figure 5C-D), we used scikit-learn (Pedregosa et al., 2011) to train L2-regularised logistic regression models that predicted location at time tL from neural activity at time tN with an inverse regularisation strength of C = 1.0. We performed this analysis in crossvalidation across locations at time tN. In other words, we trained a decoder on data where the agent was in any state stN≠s to predict all locations at time tL, and we then tested this decoder in trials where the agent was in state s at time tN. We repeated this analysis across all test locations s and averaged performance across the resulting 16 folds. We did this to test whether the RNN had a generalisable representation of future location, rather than an encoding of e.g. current location and neighboring values. In Figure 5C (left), we plot the performance of decoders trained on tN = −1 to predict location at any tL ≥ 1. In Figure 5D, we trained decoders on every pair of tN ∈ [0, 5] and tL ∈ [0, 5]. We then averaged the performance across every ‘delay’ tL − tN.

To investigate the generalisation of decoders in Figure 5E, we trained a single decoder using neural activity at time tN = 1 to predict location at time tL = 3. We then applied the same decoder to neural activity at all times t′N and quantified how well it predicted location at all times t′L. This analysis was again performed in crossvalidation. A separate decoder was trained while holding out each location s1 = s, and then tested only on trials where t′N = s. Performance was averaged across all held-out locations and across 5 independently trained RNNs. Figure 5E shows the time at which the average predictive performance was highest for each t′N. Figure S5 shows the full generalisation behaviour of the decoder.

To predict the time at which the agent would be at a particular location in Figure 5C (right), we analysed every state s separately, and then averaged over all s. For each s, we identified trials where the RNN passed through s exactly once. We trained a decoder to predict the time at which s was visited and computed the test performance as a function of the true time at which s was visited. As before, all decoders were trained in crossvalidation across current agent location.

Comparisons of RNN performance and efficiency

In Figure 5F-G, we compare three classes of RNNs, which were trained on either the reward landscape task, the moving goal task, or the static goal task with a goal that changed between trials. The performance of each RNN was quantified in each of the three tasks as the probability of choosing the optimal first action. We also computed the average parameter magnitudes of all the networks, θ|22. Finally, we computed the average firing rate of each network in the static goal task as Eτ∼π[∑t|rt|22]. All other RNN analyses apart from Figure S7 used networks trained on the reward landscape task.

Subspace identification

Prior work has investigated the extent to which working memory subspaces are orthogonal (Xie et al., 2022; Dorrell et al., 2024). However, we are primarily interested in the dynamics between subspaces, which are easier to interpret in an orthonormal coordinate system. We therefore asked whether there exists a set of orthogonal subspaces that predict the location of the agent at every time in the future. To do so, we first simulated 6,000,000 trials and collected pairs of neural activity at time t and location at t′(δ) separately for each δ. We did this in two different ways. To estimate ‘planning’ subspaces, subspace δ was defined as a decoder that predicts location at time t′(δ) = δ from neural activity at times t ∈ {− 2, −1}. To estimate ‘execution’ subspaces, subspace δ was defined as a decoder that predicts location at t′(δ) = t + δ from neural activity at any t ≥ 0.

We then defined a predictive distribution parametrised by ϕδ={Cδ∈RN×Nrec,bδ∈RN} for each δ: (34) pϕδ(st′=si)∝exp⁡(cδirt+bδi). Finally, we minimised an objective function that combines the accuracy across δs and the overlap between subspaces: (35) L(θ)=∑δ[Ert,st′(δ)∼D[−log⁡pϕδ(st′(δ))]+α1|ϕδ|1+α2|ϕδ|22]+αorthLorth, (36) Lorth:=∑δk,δl∑i,j(1−δijδδkδl)cδkiTcδlj. cδki indicates the parameter vector that predicts being at location si at a delay of (1−δijδδkδl)=0 if both i = j and δk = δl, and 1 otherwise. The parameters were optimised using ADAM with a learning rate of 5 × 10−3 until convergence or for a maximum of 2000 iterations. We used α1 = 10−4, α2 = 10−3, and αorth was annealed from 0 to 2 × 10−3 over 500 iterations. These hyperparameters were chosen because they resulted in a good approximation to the ‘true’ subspaces in the handcrafted STA.

For the analyses in Figure S6, we considered neural activity from our 6,000,000 simulated trials separately for each time-within-trial. We then performed PCA on neural activity across all trials for each time point. We analysed both how much variance was explained by the top 80 PCs, and how much variance was explained by each of the five ‘planning’ and ‘execution’ subspaces defined by ĉδi for δ = [0, 1, 2, 3, 4].

Estimating effective connectivity

To compute the effective connectivity between representations of different points in spacetime, we projected the learned network parameters into a coordinate system defined by the parameter vectors C∈RNT×Nrec. Each row of C is a normalised vector c^δi that predicts a particular point in spacetime. In this coordinate system, the input weights are given by Wineff=CWin; the output weights by Wouteff=WoutCT; and the in recurrent weights by Wreceft=CWrecCT. To analyse weights between ‘adjacent’ subspaces, we averaged the blocks of Wreceff that corresponded to weights from any subspace δ to δ + 1 and from any subspace δ to δ − 1. To avoid our analyses being biased by the fact that the networks were trained with supervised learning to predict optimal locations that could only be adjacent, we did not include the weights between subspaces 0 and 1 in this average.

To quantify the similarity between the recurrent weights in this spacetime coordinate system and different order adjacency matrices for the environment, we computed point-biserial correlations. The Δth order adjacency matrix AΔ ∈ ℝN×N was defined as a binary matrix with elements equal to 1 for pairs of states that can be reached from one another in Δ actions, and 0 for all other pairs of states. The 0th order adjacency matrix was defined as the identity matrix.

Perturbation analyses

For the perturbation analyses in Figure 6F-H, we constructed an environment where the reward function had elements (i) Rts = 1.0 for (t, s) ∈ {(0, 0), (1, 1), (2, 2), (3, 6), (4, 10), (5, 10), (6, 10)}, (ii) Rts = 0.7 for (t, s) ∈ {(1, 4), (2, 8), (3, 9)}, and Rts = 1.0 for all other points in spacetime. The RNN reliably converged to a representation of the optimal path. We first ran the network dynamics for 10 environment iterations without perturbation after the end of the normal planning period. We then continued to run the network dynamics for 10 environment iterations with a bias term defined by brecstim=brec+αc^2,8, where α is the stimulation strength. The perturbed representation was defined as the representation at the end of this perturbation period. The two example representations in Figure 6G used α = 0.3 (‘weak’) and α = 10 (‘strong’). The quantification in Figure 6G used a range of α from 0 to 10 for the RNN, and from 0 to 500 for the handcrafted STA. The control analyses (grey lines in Figure 6G and Figure S9C) were performed by running the same analysis on the RNNs, but with stimulation of the same magnitude in a random direction in neural state space. For the analysis in Figure 6H, we also removed the perturbation and ran the network dynamics for 10 environment iterations after the end of the perturbation period. All analyses in the main text focused on representations in the space of implied future trajectories, pϕδ(st′=si). The ‘representational change’ was quantified as the L1 norm of the difference from the spacetime representation at the end of the normal planning period. See Figure S9 for an analysis of the raw firing rates.

RNNs trained in changing environments

For the analyses in Figure 7, we trained another set of RNNs in a version of the reward landscape task where the transition structure changed between trials. The locations of all walls in the environment were provided as a binary input to the agent. For the performance comparison in Figure 7B (top), we evaluated the performance of these networks in the environments that the ‘single structure’ networks had been trained on.

The effective connectivity in Figure 7C-D was computed as in the RNNs trained on a single structure. For these analyses, we identified the subspaces from 1,000,000 trials in a single environment, and we repeated this analysis across 30 different environments. We computed similarities between (i) the effective connectivity estimated in an environment and the adjacency matrix of the same environment, and (ii) the effective connectivity and the adjacency matrix from a different control environment. In Figure 7E, the ‘subspace similarity’ was defined as the correlation between the set of vectors that define the subspaces, averaged over all points in spacetime. We computed the similarity between (i) subspaces identified from two sets of independent trials from the same maze, and (ii) the same number of trials from two different mazes. Recall that the effective recurrent weights between subspaces are given by Wreceft=CWrecC1. By using different subspaces in different environments, the RNN changes C between environments, which changes the effective connectivity between pairs of future subspaces (Figure 7E).

For the analyses in Figure 7G-H, we repeated the subspace identification procedure, but with two notable differences. First, we used trials across many environments to find generalised directions in neural state space that predict the future in any environment. Second, we defined the objective function in terms of future transitions τδij instead of locations. Figure 7G quantifies the effective connectivity between future transitions in consecutive subspaces (τδij and τδ+1kl) that are either ‘consistent’ (k = j), ‘adjacent’ (j≠ k but si and sk are adjacent in an environment with no walls), or any other transitions. In Figure 7H, we first projected the input corresponding to a given wall location wij onto each subspace and normalised the projection within each subspace. We then calculated the dot product between this projection and the normalised directions in neural state space that predict either (i) transitions between si and sj, (ii) transitions from si or sj to some other state sk, or (iii) transitions that do not originate at si or sj. Projection magnitudes were averaged over transitions within each of these three groups, then across subspaces, and finally the mean and standard deviation were computed across 5 RNNs.